---
title: "Python-Free Sentence-BERT in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python-Free Sentence-BERT in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  out.width = "100%"
)
```

`sbert` runs a genuine Sentence-BERT model without Python. It combines the
official Hugging Face tokenizer configuration, native ONNX Runtime inference,
attention-mask-aware mean pooling, and L2 normalization.

Downloads are never triggered by package installation, loading, examples,
tests, or this vignette. Install the native runtime and pinned model explicitly:

```{r install, eval=FALSE}
library(sbert)
install_runtime()
model_download()
```

Load the verified files and encode text:

```{r encode, eval=FALSE}
model <- load_model()
sentences <- c(
  "A student is reading a research paper.",
  "A learner studies an academic article.",
  "The bicycle is parked beside the building."
)
embeddings <- encode(sentences, model)
dim(embeddings)
topic_similarity(embeddings)
```

The output has one row per sentence and 384 columns. Normalized vectors have
unit L2 norm, so their matrix product is cosine topic_similarity.

The post-processing functions are independently usable and require no model
download. This small example masks the third token of each sentence:

```{r pooling}
token_embeddings <- array(seq_len(12), dim = c(2, 3, 2))
attention_mask <- matrix(c(1, 1, 0, 1, 1, 0), nrow = 2, byrow = TRUE)
sentence_embeddings <- sbert::pool(
  token_embeddings,
  attention_mask
)
sentence_embeddings
rowSums(sentence_embeddings^2)
sbert::topic_similarity(sentence_embeddings)
```

Use `model_status()` to inspect artifact validity,
`cache_size()` to measure disk use, and `model_remove()` to remove
the pinned model.

## Semantic topic modeling

`topics()` clusters normalized sentence embeddings with deterministic
k-means. It then identifies documents nearest each semantic center and computes
class-based TF-IDF terms from the original text. The topic count is always
explicit.

The same function accepts precomputed embeddings, so this complete example is
offline-safe:

```{r topics}
topic_text <- c(
  "Cats chase mice and sleep",
  "Dogs chase balls and sleep",
  "Neural networks learn representations",
  "Machine learning models learn patterns",
  "Bicycles have wheels and pedals",
  "Cars have wheels and engines"
)
topic_embeddings <- rbind(
  c(1, 0, 0), c(0.95, 0.05, 0),
  c(0, 1, 0), c(0.05, 0.95, 0),
  c(0, 0, 1), c(0.05, 0, 0.95)
)
topic_model <- sbert::topics(
  topic_text,
  n_topics = 3,
  embeddings = topic_embeddings,
  n_terms = 4,
  n_representatives = 1,
  keep_embeddings = TRUE
)
topic_model$topics
topic_model$terms
topic_model$representatives
```

This is document-level embedding topic discovery. It is not a probabilistic
LDA model and does not estimate per-document topic mixtures.

The term weighting follows the class-based TF-IDF of Mendonca and Figueira
(2025). Set `weighting = "bm25"` or `reduce_frequent_words = TRUE` to switch to
the BM25 and square-root variants that more aggressively suppress terms shared
across topics.

## Evaluating topic quality

Topic solutions are judged, not assumed. `coherence()` scores how often a
topic's top terms co-occur in the same documents; `topic_diversity()` reports
how distinct the topics' vocabularies are; and `summary()` combines both into a
single report with a tidy per-topic quality table.

```{r evaluate}
sbert::coherence(topic_model, measure = "npmi")
sbert::topic_diversity(topic_model)
summary(topic_model)
```

`"umass"` coherence (Mimno et al. 2011) and `"npmi"` coherence are both
computed on the input corpus, so no external reference corpus is required.

## Visualizing a topic model

Three deterministic base-graphics views summarize a model. The document map
projects the embeddings to two dimensions with classical multidimensional
scaling on cosine distance -- a reproducible alternative to UMAP.

```{r plot-sizes}
plot(topic_model, type = "sizes")
```

```{r plot-terms}
plot(topic_model, type = "terms")
```

```{r plot-map}
plot(topic_model, type = "map")
```

