---
title: "Topic Modeling: A Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Topic Modeling: A Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(sbert)
fixture <- readRDS(system.file("extdata", "feedback_embeddings.rds", package = "sbert"))
```

This tutorial fits a topic model end to end on real data and reads the result.
Every step is one verb with named arguments returning a data frame you print
directly.

## The data

`feedback_translations` ships with the package: 8,757 AI-generated mathematics
feedback messages from the Levebee application, each paired with an English
translation. The source languages include Czech, Slovak, Polish, German,
Hungarian, Romanian, Ukrainian, Russian, Mongolian, and Vietnamese.

```{r data}
head(feedback_translations)
```

The messages are short instructional hints, and some repeat verbatim — "Try
again." is the most common, appearing 11 times. That matters for topic
modeling, because a heavily repeated template would otherwise pull a topic
toward itself. `dedupe()` collapses repeats and records how often each one
occurred:

```{r dedupe, eval=FALSE}
corpus <- dedupe(feedback_translations$translation)
head(corpus)
```

## Embedding

Encoding is the one expensive step, so do it once and reuse the matrix:

```{r encode, eval=FALSE}
install_runtime()
model_download()

embeddings <- encode(corpus$text, batch_size = 64)
```

To keep this vignette buildable without a download, the remaining chunks use a
precomputed matrix for 600 of those messages, encoded with the pinned
`all-MiniLM-L6-v2`:

```{r fixture}
text <- fixture$text
embeddings <- fixture$embeddings

length(text)
dim(embeddings)
```

## Choosing how many topics

There is no correct topic count. `select_topics()` fits one model per
candidate and reports the numbers that justify a choice — so you pick from a
table rather than by habit:

```{r sweep}
sweep <- select_topics(
  text,
  n_topics = c(4, 6, 8, 10, 12),
  embeddings = embeddings,
  measure = "npmi"
)

sweep
```

Read it this way: coherence measures whether a topic's top terms actually
co-occur, diversity whether topics repeat each other's vocabulary, and
`explained` how much embedding variance sits between topics rather than
within them. Coherence usually falls as counts rise while `explained` always
rises, so the useful signal is the count *after which coherence stops
improving* — not a global maximum.

```{r sweep-plot}
plot(sweep)
```

The sweep keeps every model it fitted, so committing to one costs nothing:

```{r pick}
fit <- fitted(sweep, n_topics = 6)
fit
```

## Reading the topics

`summary()` is the report: sizes, quality, and the terms that define each
topic.

```{r summary}
summary(fit)
```

Terms are ranked by class-based TF-IDF, which favours words that separate this
topic from the others:

```{r terms}
head(terms(fit, n = 5), 15)
```

Distinctive is not the same as frequent. To see the words a topic simply uses
most — its raw \(p(term \mid topic)\) — sort by `beta` instead:

```{r terms-beta}
head(terms(fit, n = 5, sort_by = "beta"), 15)
```

Term settings can be changed without refitting anything, because terms depend
only on the fitted assignments and the text:

```{r terms-retune}
head(terms(fit, n = 5, weighting = "bm25"), 10)
```

If a word appears in nearly every document it defines nothing. Exclude it and
look again:

```{r stopwords}
head(terms(fit, n = 5, stop_words = stop_words(add = c("try", "click"))), 10)
```

## Evidence for each topic

Terms describe a topic; documents prove it. `representatives()` ranks by
margin — how much more a document belongs to its own topic than to its closest
rival — which surfaces the clear cases rather than merely the central ones:

```{r representatives}
head(representatives(fit, n = 2), 8)
```

## Quality

```{r quality}
coherence(fit, measure = "npmi")
```

```{r diversity}
topic_diversity(fit)
```

Diversity is the share of distinct terms across all topics' term lists: 1
means no topic repeats another's vocabulary, low values mean the model has
split one theme several ways.

## Which topics are neighbours?

```{r hierarchy}
tree <- topic_hierarchy(fit)
tree
```

```{r hierarchy-plot}
plot(tree)
```

Merging down the tree is not the same as refitting at a smaller count: it
keeps the finer structure as a nesting, which is usually what you want for
reporting.

```{r reduce}
smaller <- reduce_topics(fit, n_topics = 5)
smaller
```

## Assigning new messages

```{r predict, eval=FALSE}
predict(fit, c("Well done, that is correct!", "Look at the picture again."))
```

Hard assignment gives each document one topic. When you need the full picture
— how strongly every document belongs to every topic — ask for membership:

```{r membership}
head(topic_membership(fit), 8)
```

Probabilities sum to 1 within a document, and `rank` 1 is its strongest topic.

## Visual summary

`plot()` renders a fitted model several ways. Sizes counts documents per topic:

```{r plot-sizes, fig.height=5}
plot(fit, type = "sizes")
```

`type = "terms"` defaults to the class-based TF-IDF terms. The `by` argument
switches the ranking to the raw within-topic count (`"frequency"`) or the
generative probability (`"beta"`), and accepts several metrics at once — one
column each — so the *says-most* and *says-distinctively* views sit side by
side (shown here for one topic):

```{r plot-terms, fig.height=6}
plot(fit, type = "terms")
```

```{r plot-terms-by, fig.height=3.5}
plot(fit, type = "terms", by = c("frequency", "score", "beta"), topics = 1)
```

`type = "representatives"` lists each topic's centroid-nearest messages — the
evidence that a label means what it claims. Ask for more than the model stored
and they are recomputed from the retained embeddings:

```{r plot-representatives, fig.height=6}
plot(fit, type = "representatives", n_representatives = 8)
```

`type = "fit"` is the whole per-topic report in one figure — all three term
views beside the representative documents, one row per topic. `per_topic = TRUE`
gives each topic its own figure with the documents stacked beneath the terms
(shown here for a single topic):

```{r plot-fit, fig.height=4}
plot(fit, type = "fit", per_topic = TRUE, topics = 1)
```

## Guided topics

When you already know what you are looking for, seed it. Seeds become the
starting centroids and name their topics:

```{r seeds, eval=FALSE}
guided <- topics(
  text,
  n_topics = 4,
  embeddings = embeddings,
  seeds = c(
    praise = "well done excellent correct answer",
    retry = "try again check your answer",
    instruction = "read the task and choose the picture",
    counting = "count the numbers and add them"
  )
)
```

With `fixed_seeds = TRUE` the centroids are frozen and every document is
assigned to its nearest seed — zero-shot classification into categories you
defined.

## Where to go next

- `keywords()` extracts keywords from documents directly, without a topic
  model.
- `segment()` splits documents into sentences or clauses, and `blend()`
  carries each segment's parent-document context into its embedding.
- `?topics` documents every argument used above.
