---
title: "Genuine Sentence-BERT in R Without Python"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
if (!"package:sbert" %in% search()) {
  pkgload::load_all(quiet = TRUE)
}

model_source <- Sys.getenv("SBERT_TEST_MODEL_DIR", unset = "")
if (!nzchar(model_source)) {
  stop(paste0(
    "Set SBERT_TEST_MODEL_DIR to a directory containing model.onnx and ",
    "tokenizer.json before rendering."
  ))
}
cache_dir <- tempfile("sbert-html-cache-")
model_directory <- sbert:::sbert_model_directory(cache_dir)
dir.create(model_directory, recursive = TRUE)
copied <- file.copy(
  file.path(model_source, c("model.onnx", "tokenizer.json")),
  model_directory
)
stopifnot(all(copied))
```

This document executes the official `all-MiniLM-L6-v2` model through native
ONNX Runtime. Python is not installed, imported, or called by the R pipeline.

## Load the verified model

```{r model}
model_status <- sbert_model_status(cache_dir)
model_status[, c("file", "exists", "valid", "actual_bytes")]

model <- sbert_load_model(cache_dir, backend = "cpu", threads = 1L)
model
```

## Encode sentences

```{r encode}
sentences <- c(
  "A student is reading a research paper.",
  "A learner studies an academic article.",
  "The bicycle is parked beside the building.",
  "A bike stands next to a structure."
)

embedding_matrix <- sbert_encode(sentences, model, batch_size = 2L)
str(embedding_matrix)
dim(embedding_matrix)
summary(as.numeric(embedding_matrix))
rowSums(embedding_matrix^2)
anyNA(embedding_matrix)
```

Each sentence becomes one 384-dimensional, L2-normalized vector.

## Cosine similarity

```{r similarity}
similarity_matrix <- sbert_similarity(embedding_matrix)
round(similarity_matrix, 3)

within_topic_similarity <- mean(c(
  similarity_matrix[1, 2],
  similarity_matrix[3, 4]
))
between_topic_similarity <- mean(similarity_matrix[1:2, 3:4])

data.frame(
  comparison = c("Semantically related pairs", "Between-topic pairs"),
  mean_cosine_similarity = c(
    within_topic_similarity,
    between_topic_similarity
  )
)
```

The semantically related pairs should be more similar than the cross-topic
pairs. The test suite also compares every embedding value against the official
SentenceTransformers reference with a maximum permitted absolute error of
`1e-6`.
