---
title: "Sentence Embeddings in R with Existing Packages"
subtitle: "An offline, R-native fallback when Sentence-BERT cannot run"
date: "2026-07-17"
output:
  html_document:
    toc: true
    self_contained: true
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  error = FALSE,
  message = TRUE,
  warning = TRUE
)

required_packages <- c(
  "knitr",
  "quanteda",
  "reticulate",
  "RSpectra"
)

package_available <- vapply(
  required_packages,
  requireNamespace,
  logical(1),
  quietly = TRUE
)

stopifnot(all(package_available))

package_summary <- data.frame(
  package = required_packages,
  version = vapply(
    required_packages,
    function(package_name) {
      as.character(utils::packageVersion(package_name))
    },
    character(1)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(package_summary, caption = "Packages used by this report")
```

## What this example proves

R can perform a complete sentence-embedding workflow locally: prepare text,
produce one numeric vector per sentence, and calculate cosine topic_similarity. The
installed packages can do this with **Latent Semantic Analysis (LSA)**.

This rendered run is **not Sentence-BERT**. Sentence-BERT requires a compatible
transformer runtime, correct tokenizer, pretrained weights, pooling, and
normalization. The active environment has `reticulate`, but not the Python
`sentence_transformers` runtime.

```{r backend-audit}
sbert_available <- tryCatch(
  reticulate::py_module_available("sentence_transformers"),
  error = function(error_condition) {
    warning(sprintf(
      "Could not inspect the Python environment: %s",
      conditionMessage(error_condition)
    ))
    FALSE
  }
)

backend_summary <- data.frame(
  component = c("R-to-Python interface", "Sentence-BERT runtime"),
  available = c(TRUE, sbert_available),
  implication = c(
    "reticulate is installed",
    if (sbert_available) {
      "Actual Sentence-BERT inference can be configured"
    } else {
      "This report uses the local LSA fallback"
    }
  ),
  stringsAsFactors = FALSE
)

knitr::kable(backend_summary, caption = "Runtime audit")
```

## Synthetic sentences

The example contains two topics. Each topic has two sentences that use
different vocabulary and a third sentence that bridges the vocabulary. This
small construction lets LSA learn the intended latent grouping.

```{r synthetic-data}
sentence_data <- data.frame(
  document_id = c(
    "cat_1",
    "cat_2",
    "cat_bridge",
    "car_1",
    "car_2",
    "car_bridge"
  ),
  topic = rep(c("animals", "vehicles"), each = 3L),
  text = c(
    "The cat sleeps on the sofa.",
    "A kitten naps on the couch.",
    "The cat and kitten sleep together on the sofa and couch.",
    "The car needs a new engine.",
    "The automobile requires motor repair.",
    "The car and automobile have an engine and motor."
  ),
  stringsAsFactors = FALSE
)

str(sentence_data)
head(sentence_data)
summary(sentence_data)
vapply(sentence_data, class, character(1))

stopifnot(
  nrow(sentence_data) > 1L,
  !anyNA(sentence_data),
  all(nzchar(sentence_data$document_id)),
  all(nzchar(sentence_data$text)),
  anyDuplicated(sentence_data$document_id) == 0L,
  anyDuplicated(sentence_data$text) == 0L
)

knitr::kable(
  sentence_data[, c("document_id", "topic", "text")],
  caption = "Synthetic input data"
)
```

## Create local LSA embeddings

The pipeline tokenizes the sentences, removes English stop words, stems the
remaining words, creates a TF--IDF matrix, and reduces it to two latent
dimensions with truncated singular-value decomposition.

```{r lsa-embeddings}
text_corpus <- quanteda::corpus(
  sentence_data,
  text_field = "text",
  docid_field = "document_id"
)

text_tokens <- quanteda::tokens(text_corpus, remove_punct = TRUE) |>
  quanteda::tokens_tolower() |>
  quanteda::tokens_remove(quanteda::stopwords("en")) |>
  quanteda::tokens_wordstem(language = "english")

token_counts <- quanteda::ntoken(text_tokens)
stopifnot(!anyNA(token_counts), all(token_counts > 0L))

text_dfm <- quanteda::dfm(text_tokens) |>
  quanteda::dfm_tfidf(scheme_tf = "count", scheme_df = "inverse")

text_matrix <- methods::as(text_dfm, "dgCMatrix")
latent_dimensions <- 2L

stopifnot(
  identical(nrow(text_matrix), nrow(sentence_data)),
  ncol(text_matrix) > latent_dimensions,
  latent_dimensions < min(dim(text_matrix)),
  !anyNA(text_matrix@x),
  all(is.finite(text_matrix@x))
)

text_svd <- tryCatch(
  RSpectra::svds(text_matrix, k = latent_dimensions),
  error = function(error_condition) {
    stop(sprintf(
      "LSA decomposition failed: %s",
      conditionMessage(error_condition)
    ))
  }
)

str(text_svd)
names(text_svd)
dim(text_svd$u)
length(text_svd$d)
dim(text_svd$v)

embedding_matrix <- sweep(
  text_svd$u,
  MARGIN = 2L,
  STATS = text_svd$d,
  FUN = "*"
)

rownames(embedding_matrix) <- quanteda::docnames(text_dfm)
colnames(embedding_matrix) <- sprintf(
  "lsa_%02d",
  seq_len(ncol(embedding_matrix))
)

embedding_row_norms <- sqrt(rowSums(embedding_matrix^2))

stopifnot(
  is.matrix(embedding_matrix),
  identical(dim(embedding_matrix), c(nrow(sentence_data), latent_dimensions)),
  !anyNA(embedding_matrix),
  all(is.finite(embedding_matrix)),
  all(embedding_row_norms > 0)
)

str(embedding_matrix)
head(embedding_matrix)
dim(embedding_matrix)
summary(as.vector(embedding_matrix))

knitr::kable(
  round(embedding_matrix, 4),
  caption = "Two-dimensional LSA sentence embeddings (not Sentence-BERT)"
)
```

## Cosine topic_similarity

After row normalization, the matrix product gives every pairwise cosine
topic_similarity. Values near 1 indicate vectors pointing in similar directions;
values near 0 indicate little shared direction.

```{r cosine-topic_similarity}
normalized_embedding_matrix <- embedding_matrix / embedding_row_norms
similarity_matrix <- tcrossprod(normalized_embedding_matrix)

dimnames(similarity_matrix) <- list(
  sentence_data$document_id,
  sentence_data$document_id
)

same_topic_pairs <- outer(
  sentence_data$topic,
  sentence_data$topic,
  FUN = "=="
) & upper.tri(similarity_matrix)

different_topic_pairs <- outer(
  sentence_data$topic,
  sentence_data$topic,
  FUN = "!="
) & upper.tri(similarity_matrix)

within_topic_similarity <- mean(similarity_matrix[same_topic_pairs])
between_topic_similarity <- mean(similarity_matrix[different_topic_pairs])

stopifnot(
  is.matrix(similarity_matrix),
  identical(dim(similarity_matrix), c(nrow(sentence_data), nrow(sentence_data))),
  !anyNA(similarity_matrix),
  all(is.finite(similarity_matrix)),
  isTRUE(all.equal(
    similarity_matrix,
    t(similarity_matrix),
    tolerance = 1e-12
  )),
  isTRUE(all.equal(
    unname(diag(similarity_matrix)),
    rep(1, nrow(sentence_data)),
    tolerance = 1e-12
  )),
  within_topic_similarity > between_topic_similarity
)

str(similarity_matrix)
head(similarity_matrix)
dim(similarity_matrix)
summary(as.vector(similarity_matrix))

knitr::kable(
  round(similarity_matrix, 3),
  caption = "Cosine topic_similarity between LSA embeddings"
)
```

```{r verification-summary}
verification_summary <- data.frame(
  check = c(
    "Missing input values",
    "Duplicate document IDs",
    "TF--IDF dimensions",
    "Embedding dimensions",
    "Similarity dimensions",
    "Mean within-topic topic_similarity",
    "Mean between-topic topic_similarity"
  ),
  result = c(
    as.character(anyNA(sentence_data)),
    as.character(anyDuplicated(sentence_data$document_id)),
    paste(dim(text_matrix), collapse = " x "),
    paste(dim(embedding_matrix), collapse = " x "),
    paste(dim(similarity_matrix), collapse = " x "),
    sprintf("%.3f", within_topic_similarity),
    sprintf("%.3f", between_topic_similarity)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(verification_summary, caption = "Pipeline verification")
```

## Interpretation and limitation

The same-topic sentences are closer than the different-topic sentences in this
deliberately small example. This verifies the mechanics of an embedding and
topic_similarity workflow using only packages already installed.

It does **not** establish equivalence with Sentence-BERT. LSA is fitted to the
current corpus, largely ignores word order and context, and can connect words
only through patterns present in that corpus. Sentence-BERT instead supplies
pretrained, contextual sentence representations. Genuine Sentence-BERT output
should replace `embedding_matrix` once a transformer runtime or authenticated
embedding service is available; the cosine-topic_similarity and verification steps
can remain the same.
