---
title: "Python-Free Sentence-BERT Topic Modeling in R"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
if (!"package:sbert" %in% search()) {
  pkgload::load_all(quiet = TRUE)
}

model_source <- Sys.getenv("SBERT_TEST_MODEL_DIR", unset = "")
if (!nzchar(model_source)) {
  stop(paste0(
    "Set SBERT_TEST_MODEL_DIR to a directory containing model.onnx and ",
    "tokenizer.json before rendering."
  ))
}
cache_dir <- tempfile("sbert-topic-html-cache-")
model_directory <- sbert:::sbert_model_directory(cache_dir)
dir.create(model_directory, recursive = TRUE)
copied <- file.copy(
  file.path(model_source, c("model.onnx", "tokenizer.json")),
  model_directory
)
stopifnot(all(copied))
```

This example performs document-level topic discovery with genuine
`all-MiniLM-L6-v2` embeddings. The complete R pipeline is Python-free:

1. Hugging Face-compatible tokenization with `tok`.
2. Native ONNX inference with `onnxr`.
3. Deterministic semantic clustering.
4. Representative-document and class TF-IDF topic summaries.

## Synthetic corpus

The expected labels below are used only to evaluate the discovered partition;
they are never passed to the topic model.

```{r corpus}
topic_data <- data.frame(
  expected_topic = rep(c("astronomy", "cooking", "finance"), each = 5L),
  text = c(
    "A telescope captured a distant galaxy filled with newborn stars.",
    "Astronomers measured the orbit of a planet around a nearby star.",
    "A spacecraft photographed craters across the surface of the moon.",
    "The observatory detected light from a supernova in deep space.",
    "A satellite mapped clouds of dust throughout the Milky Way galaxy.",
    "The chef sauteed garlic and onions in olive oil before adding tomatoes.",
    "Bread dough rises with yeast before it is baked in the oven.",
    "The recipe says to whisk eggs and sugar into the cake batter.",
    "Vegetables and spices simmered in broth to make a hearty soup.",
    "The cook grilled salmon seasoned with fresh lemon and herbs.",
    "The central bank raised interest rates to slow inflation.",
    "Investors bought government bonds when the stock market declined.",
    "Strong quarterly earnings sent the company share price higher.",
    "The borrower compared mortgage loans and their monthly payments.",
    "The portfolio spreads risk among equities, bonds, and cash."
  ),
  stringsAsFactors = FALSE
)

str(topic_data)
table(topic_data$expected_topic)
anyNA(topic_data)
anyDuplicated(topic_data$text)
```

## Fit the topic model

```{r fit}
model <- sbert_load_model(cache_dir, backend = "cpu", threads = 1L)

topic_model <- sbert_topics(
  text = topic_data$text,
  n_topics = 3L,
  model = model,
  batch_size = 8L,
  n_terms = 6L,
  n_representatives = 2L,
  keep_embeddings = TRUE
)

topic_model
str(topic_model, max.level = 1L)
dim(topic_model$embeddings)
summary(as.numeric(topic_model$embeddings))
rowSums(topic_model$embeddings^2)
```

## Topic summaries

```{r summaries}
topic_model$topics
topic_model$terms
topic_model$representatives[, c("topic", "rank", "text", "distance")]
```

Topic numbers are arbitrary cluster identifiers. Interpretation should use the
terms and representative documents rather than assuming that topic 1 has a
predefined meaning.

## Label-invariant evaluation

```{r evaluation}
assignment_table <- table(
  expected = topic_data$expected_topic,
  discovered = topic_model$documents$topic
)
assignment_table

true_same_topic <- outer(
  topic_data$expected_topic,
  topic_data$expected_topic,
  `==`
)
discovered_same_topic <- outer(
  topic_model$documents$topic,
  topic_model$documents$topic,
  `==`
)
upper_triangle <- upper.tri(true_same_topic)
rand_agreement <- mean(
  true_same_topic[upper_triangle] ==
    discovered_same_topic[upper_triangle]
)

data.frame(
  metric = c("Pairwise Rand agreement", "Number of discovered topics"),
  value = c(rand_agreement, length(unique(topic_model$documents$topic)))
)
```

This corpus has perfect label-invariant agreement (`1.0`): every document from
the same expected domain was placed together, and all three domains remained
separate.
