---
title: "Semantic Topics in Levebee AI Mathematics Feedback — an sbert Tutorial"
author: "sbert 0.5.2"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
library(sbert)
```

# Why this tutorial

The `feedback_translations` dataset bundled with **sbert** contains 8,757
AI-generated feedback messages from the Levebee mathematics application,
translated into English from ten source languages. Nobody can read nearly nine
thousand messages, and many repeat verbatim — "Try again." is the most common,
appearing 11 times — so simple frequency lists tell you about templates,
not content. Topic modeling answers the question the raw data cannot: *what
kinds of feedback does the system actually give, and in what proportions?*

The answer has to survive scrutiny, so every step here is deterministic:
rerunning this document reproduces every number, bar, and sentence exactly.
The tutorial builds one six-topic model and then inspects it from three
deliberately different angles:

1. **Model quality** — coherence and diversity, so you know which topics to
   trust before interpreting any of them.
2. **Two keyword views per topic** — raw within-topic counts (what the topic
   *says most*) against class-based TF-IDF (what the topic *alone* says).
   These are different lists by construction, and the disagreement between
   them is itself informative.
3. **Representative sentences** — the centroid-nearest messages, which are
   the auditable evidence that a topic label means what it claims.

| Component | Verb | Question it answers |
|---|---|---|
| Topic model | `topics()` | What groups exist, and how big are they? |
| Evaluation | `coherence()`, `topic_diversity()` | Which topics are trustworthy? |
| Frequent keywords | `terms(sort_by = "beta")` | What does each topic talk about most? |
| Distinctive keywords | `topic_model$terms` | What does each topic talk about that others do not? |
| Evidence | `topic_model$representatives` | Do real messages support the label? |

```{r library, eval=FALSE}
library(sbert)
```

Fourteen revision-pinned models are available; this tutorial uses the
default, `all-MiniLM-L6-v2` (the field's standard quality-per-megabyte
English embedder). The menu, with each model's dimension, input limit,
language coverage, and download size:

```{r menu}
models()
```

# The corpus, deduplicated

Embedding the same string twice wastes computation and — more importantly —
lets repeated templates skew the clustering geometry: every copy of a
message like "Try again." acts as another point pulling on the same
centroid. `dedupe()` therefore collapses the corpus to its
distinct non-blank messages (each kind votes once) while keeping the row
frequencies, which return as weights when sizes are reported.

```{r data}
corpus <- dedupe(feedback_translations$translation)
nrow(corpus)


```

Three explicit steps turn the corpus into numbers, and each exists for a
reason. Called without arguments, every model function uses the package
default, `all-MiniLM-L6-v2` — a 6-layer distilled English model that is the
field's standard quality-per-megabyte starter (`models()` lists the
other thirteen pinned options). `model_download()` is the only step that
touches the network: it fetches the model's ONNX graph (90.4 MB) and
tokenizer into the local cache, locked to one immutable Hugging Face
revision and refused
unless every byte matches the SHA-256 hashes pinned inside the package — so
the model you run today is provably the model you run next year. Once the
files are installed the call is a no-op. `load_model()` then builds
the in-memory model from those verified files: the Rust tokenizer, the ONNX
inference session, and the model's own settings (384 dimensions, 256-token
limit, mean pooling).

`encode()` does the actual work: each message is tokenized into word
pieces, run through the network (one vector per token), mean-pooled into one
vector per message with padding masked out, and L2-normalized. The
normalization is what makes everything downstream cheap — on unit vectors,
cosine similarity is a plain dot product, which is exactly what the k-means
clustering and the centroid distances in the panels below rely on. The
result is an 8,005 × 384 matrix, numerically identical to Python
`SentenceTransformers` output and bit-identical on every rerun — which is
why this tutorial recomputes it (about ten seconds) instead of shipping a
cache file.

```{r embeddings}
model_download()
minilm <- load_model()
embeddings <- encode(corpus$text, model = minilm)
```

# The six-topic model

Six topics is a reviewed choice for this corpus, not an algorithmic guess —
**sbert** deliberately refuses to pick `n_topics` for you, because the right
granularity depends on what the analysis is *for* (here: an editorial review
of feedback types, where six kinds are actionable and sixty are not).

```{r model}
topic_model <- topics(
  corpus$text,
  n_topics = 6,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 5
)
topic_model
```

## Should you trust these topics?

Interpretation comes *after* evaluation, because a beautifully labeled topic
with poor coherence is a story about noise. NPMI coherence asks whether a
topic's top terms actually co-occur in its messages (+1 = always together,
−1 = never); diversity asks whether topics share their vocabulary or own it.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
coherence(topic_model, measure = "npmi")
```

Carry these numbers into the per-topic panels below: the topics with the
highest NPMI are the rigid instructional templates (regular phrasing is
exactly what co-occurrence statistics reward), and any topic with visibly
lower coherence should be read through its representative sentences rather
than its keywords.

## Size on two scales

The model counts *distinct* messages, but the application sent some messages
hundreds of times, so editorial priority follows the weighted share, not the
distinct share. `topic_sizes()` reports both scales in one call; the
gap between `proportion` and `weighted_share` measures how template-driven
each topic is — a topic whose weighted share far exceeds its distinct share
is a small repertoire of heavily reused messages.

```{r sizes}
plot(topic_model, type = "sizes")
```

```{r weighted}
topic_sizes(topic_model, weights = corpus$n)
```

# Two keyword views, one topic

`terms(sort_by = "beta")` returns the empirical probability of each word given the topic
— the *generative* view, dominated by whatever the topic says most often.
The class-based TF-IDF scores in `topic_model$terms` are the *discriminative*
view: they promote words this topic uses and others do not, and demote the
corpus-wide vocabulary ("picture", "choose") that every feedback message
shares. Neither list is "the" keywords; a topic is characterized by the pair.
When the two lists agree, the topic has a vocabulary of its own; when they
disagree, the counts list is telling you about the corpus and only the
TF-IDF list about the topic.

```{r beta}
beta <- terms(topic_model, n = NULL, sort_by = "beta")
head(subset(beta, topic == 1), 8)
```

```{r terms}
head(subset(topic_model$terms, topic == 1), 8)
```

# The topics at a glance

The two-panel drawing this tutorial once carried by hand is now built into the
package. `plot(topic_model, type = "terms")` shows each topic's distinctive
class-based TF-IDF terms, each bar annotated with its score:

```{r plot-terms, fig.width = 9, fig.height = 7}
plot(topic_model, type = "terms")
```

`type = "fit"` is the whole per-topic report in one figure: for every topic
(one row) it draws all three keyword views — raw within-topic count, class-based
TF-IDF, and generative probability (beta) — followed by the topic's
centroid-nearest messages with their cosine similarity to the centroid. Reading
across a row shows a label, why it is distinctive, and the evidence that
supports it.

```{r plot-fit, fig.width = 13, fig.height = 14}
plot(topic_model, type = "fit", n_terms = 8, n_representatives = 10)
```
The document text is clipped to fit; for the full, untruncated evidence behind
each topic, ask for the representatives directly:

```{r representatives}
subset(topic_model$representatives, rank <= 2)[
  , c("topic", "rank", "text", "distance")
]
```

# Where to go from here

The model built here is reusable, not just describable. `predict()` assigns
any new feedback message to these six topics without refitting;
`topic_membership()` replaces the hard assignment with graded probabilities
when a message sits between topics; and `topic_gamma()` combined with
`segment()` shows when a single multi-sentence message spans several
feedback types. For multilingual work on the *source* messages (the
`feedback` column spans ten languages), change one argument —
`encode(corpus$text, model = "paraphrase-multilingual-MiniLM-L12-v2")`
— and the rest of this document runs unchanged.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```