---
title: "Feedback Translation Topic Model"
subtitle: "Python-free Sentence-BERT analysis of 8,987 feedback records"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  output_directory: "outputs/feedback_translation_topics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5,
  dpi = 120
)
project_root <- normalizePath("..", mustWork = TRUE)
artifact_directory <- file.path(project_root, params$output_directory)
analysis_result <- readRDS(
  file.path(artifact_directory, "feedback_translation_topic_model.rds")
)
assignments <- analysis_result$assignments
topics <- analysis_result$topic_summary
terms <- analysis_result$topic_terms
representatives <- analysis_result$representatives
diagnostics <- analysis_result$topic_count_diagnostics
data_quality <- analysis_result$data_quality
method <- analysis_result$method
```

<style>
body { color: #243447; }
h1, h2, h3 { color: #17365d; }
.kpi-grid { display: grid; grid-template-columns: repeat(4, minmax(150px, 1fr)); gap: 14px; margin: 16px 0 24px; }
.kpi { background: #f3f7fb; border-left: 5px solid #2d6a9f; border-radius: 6px; padding: 14px 16px; }
.kpi .value { color: #17365d; font-size: 27px; font-weight: 700; line-height: 1.1; }
.kpi .label { color: #5b6b7c; font-size: 12px; margin-top: 6px; text-transform: uppercase; letter-spacing: .04em; }
.callout { background: #fff7df; border-left: 5px solid #e3a008; border-radius: 5px; padding: 14px 18px; margin: 16px 0 22px; }
.topic-card { border: 1px solid #d9e2ec; border-radius: 7px; padding: 14px 18px; margin: 13px 0; background: #ffffff; }
.topic-title { color: #17365d; font-size: 18px; font-weight: 700; }
.topic-meta { color: #5b6b7c; margin: 4px 0 9px; }
.terms { color: #2d6a9f; font-weight: 600; }
table { font-size: 13px; }
th { background: #2d6a9f !important; color: white !important; }
@media (max-width: 800px) { .kpi-grid { grid-template-columns: repeat(2, minmax(130px, 1fr)); } }
</style>

```{r kpis, results='asis'}
kpi_values <- c(
  format(nrow(assignments), big.mark = ","),
  format(sum(assignments$included), big.mark = ","),
  format(length(unique(assignments$translation[assignments$included])), big.mark = ","),
  as.character(nrow(topics))
)
kpi_labels <- c("Source rows", "Modeled rows", "Distinct translations", "Exploratory topics")
kpi_html <- paste0(
  '<div class="kpi"><div class="value">',
  kpi_values,
  '</div><div class="label">',
  kpi_labels,
  '</div></div>'
)
cat('<div class="kpi-grid">', paste(kpi_html, collapse = ""), '</div>')
```

<div class="callout">
<strong>Interpretation note.</strong> This is an exploratory embedding-based topic model, not a validated taxonomy. Topic IDs are arbitrary; automatic labels come from ranked class-based TF-IDF terms. Cosine distance is proximity to a centroid, not a probability or confidence score. The supplied <code>translation</code> field was used as-is.
</div>

## Topic prevalence

```{r topic-size-plot}
topic_axis_label <- sprintf("T%d: %s", topics$topic, topics$topic_label)
topic_color <- ifelse(
  topics$topic %in% c(12L, 15L),
  "#d99032",
  "#2d6a9f"
)
graphics::par(mar = c(5, 15, 2, 1))
graphics::barplot(
  rev(topics$n_rows),
  names.arg = rev(topic_axis_label),
  horiz = TRUE,
  las = 1,
  col = rev(topic_color),
  border = NA,
  xlab = "Feedback rows",
  main = "Rows assigned to each topic"
)
```

The first five topics account for `r sprintf('%.1f%%', 100 * sum(topics$proportion[1:5]))` of modeled rows. This concentration reflects a template-heavy corpus: many records are variants of “between” instructions with different named objects.

```{r topic-table}
topic_display <- topics[
  ,
  c(
    "topic",
    "topic_label",
    "n_rows",
    "proportion",
    "n_unique_translations",
    "median_cosine_distance",
    "p90_cosine_distance",
    "top_terms"
  )
]
topic_display$proportion <- sprintf("%.1f%%", 100 * topic_display$proportion)
topic_display$median_cosine_distance <- sprintf(
  "%.3f",
  topic_display$median_cosine_distance
)
topic_display$p90_cosine_distance <- sprintf(
  "%.3f",
  topic_display$p90_cosine_distance
)
names(topic_display) <- c(
  "Topic",
  "Automatic label",
  "Rows",
  "Share",
  "Unique texts",
  "Median distance",
  "P90 distance",
  "Top terms"
)
knitr::kable(topic_display, format = "html", escape = TRUE)
```

## Topic profiles

```{r topic-profiles, results='asis'}
invisible(lapply(
  seq_len(nrow(topics)),
  function(topic_row) {
    topic_id <- topics$topic[[topic_row]]
    representative_rows <- representatives[
      representatives$topic == topic_id & representatives$rank <= 3L,
      c("rank", "translation", "cosine_distance_to_centroid"),
      drop = FALSE
    ]
    names(representative_rows) <- c("Rank", "Representative translation", "Cosine distance")
    representative_rows$`Cosine distance` <- sprintf(
      "%.3f",
      representative_rows$`Cosine distance`
    )
    cat(
      '<div class="topic-card"><div class="topic-title">Topic ',
      topic_id,
      ': ',
      htmltools::htmlEscape(topics$topic_label[[topic_row]]),
      '</div><div class="topic-meta">',
      format(topics$n_rows[[topic_row]], big.mark = ","),
      ' rows · ',
      sprintf("%.1f%%", 100 * topics$proportion[[topic_row]]),
      ' of modeled feedback · ',
      format(topics$n_unique_translations[[topic_row]], big.mark = ","),
      ' unique translations</div><div class="terms">',
      htmltools::htmlEscape(topics$top_terms[[topic_row]]),
      '</div><br>',
      sep = ""
    )
    print(knitr::kable(representative_rows, format = "html", escape = TRUE))
    cat('</div>')
  }
))
```

## Six-topic solution

The primary model uses six topics, as requested. Candidate counts from 4 through 16 were fitted with the same deterministic initialization and remain available in the diagnostics table for comparison. The six-topic result is a broader summary: because the corpus is dominated by repeated “between” task templates, large clusters combine several related phrasings and object variants.

```{r diagnostics-plot}
graphics::par(mar = c(5, 5, 2, 5))
graphics::plot(
  diagnostics$n_topics,
  100 * diagnostics$between_total_ratio,
  type = "b",
  pch = 19,
  col = "#2d6a9f",
  xlab = "Number of topics",
  ylab = "Between / total SS (%)",
  main = "Topic-count diagnostics"
)
graphics::points(
  diagnostics$n_topics[diagnostics$selected],
  100 * diagnostics$between_total_ratio[diagnostics$selected],
  pch = 21,
  cex = 1.8,
  bg = "#e3a008",
  col = "#8a5a00"
)
```

```{r diagnostics-table}
diagnostic_display <- diagnostics[
  ,
  c(
    "n_topics",
    "calinski_harabasz",
    "between_total_ratio",
    "normalized_entropy",
    "min_topic_size",
    "median_topic_size",
    "max_topic_size",
    "selected"
  )
]
diagnostic_display$between_total_ratio <- sprintf(
  "%.1f%%",
  100 * diagnostic_display$between_total_ratio
)
diagnostic_display$calinski_harabasz <- sprintf(
  "%.1f",
  diagnostic_display$calinski_harabasz
)
diagnostic_display$normalized_entropy <- sprintf(
  "%.3f",
  diagnostic_display$normalized_entropy
)
names(diagnostic_display) <- c(
  "Topics",
  "Calinski-Harabasz",
  "Between/total SS",
  "Normalized entropy",
  "Smallest topic",
  "Median topic",
  "Largest topic",
  "Selected"
)
knitr::kable(diagnostic_display, format = "html", escape = TRUE)
```

## Translation and data quality

Eleven blank translations were excluded from modeling but retained in the row-level output. Repeated translations were embedded once for efficiency and then mapped back to all source rows, so their observed frequency still influences topic prevalence.

The review flags are deliberately conservative screening indicators: `r sum(assignments$translation_has_cyrillic)` rows contain Cyrillic, `r sum(assignments$translation_has_non_ascii_latin)` contain accented non-ASCII Latin characters, and `r sum(assignments$source_equals_translation)` nonblank rows exactly equal the source. These categories overlap, and equality can be legitimate when the source is already English.

```{r quality-table}
quality_display <- data_quality
names(quality_display) <- c("Metric", "Value")
quality_display$Metric <- gsub("_", " ", quality_display$Metric)
knitr::kable(quality_display, format = "html", escape = TRUE)
```

### Sample rows flagged for translation review

```{r quality-examples}
flagged_examples <- utils::head(
  assignments[assignments$translation_review_flag, c(
    "row_id",
    "feedback",
    "translation",
    "translation_has_cyrillic",
    "translation_has_non_ascii_latin",
    "source_equals_translation",
    "topic"
  )],
  30L
)
names(flagged_examples) <- c(
  "Row",
  "Source feedback",
  "Supplied translation",
  "Cyrillic",
  "Accented Latin",
  "Source-equal",
  "Topic"
)
knitr::kable(flagged_examples, format = "html", escape = TRUE)
```

## Method

```{r method-table}
method_display <- method
names(method_display) <- c("Item", "Value")
method_display$Item <- gsub("_", " ", method_display$Item)
knitr::kable(method_display, format = "html", escape = TRUE)
```

The accompanying workbook contains the complete 8,987-row assignment table, ranked terms, distinct representative translations, model-selection diagnostics, repeated-text audit, and method metadata.
