---
title: "Sentence, Clause, or Phrase? Segmentation Levels Compared on MCSE"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressMessages(devtools::load_all("..", quiet = TRUE))
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# Why compare levels

The clause-level MCSE vignette settled on one segmentation; this companion
shows what the *other* two levels would have put in its place. Everything
else is held fixed — the same corpus, the same `multi-qa-MiniLM-L6-cos-v1`
embeddings, the same deterministic thirty-topic model, the same
margin-ranked selection — so any difference you see between the three
statement sets is caused by the `level` argument alone.

The trade-off each level makes:

- **Sentence** splits only at `.`, `?`, `!`. Units are complete thoughts,
  but a compound academic sentence stays whole, so evidence runs longer and
  a single unit can straddle two ideas.
- **Clause** (the default) additionally splits at `;`, `:`, spaced dashes,
  and subordinating hinges ("which", "while", "in terms of"). Structured
  headers stand apart and stacked claims come apart, while comma
  enumerations stay intact.
- **Phrase** additionally splits at commas. Units are shortest and most
  pointed, but enumerations shatter and a unit can lose the verb that made
  it a claim.

# One pipeline, three levels

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
corpus <- dedupe(paste(trimws(usable$TI), trimws(usable$AB), sep = " — "))
```

```{r model}
qa_minilm <- load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
embeddings <- encode(corpus$text, model = qa_minilm)
topic_model <- topics(
  corpus$text,
  n_topics = 30,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 3
)
```

Segmentation is the only line that changes:

```{r segment}
abstracts <- unique(trimws(usable$AB))
sentence_units <- segment(abstracts, level = "sentence", merge_below = 5)
clause_units <- segment(abstracts, level = "clause", merge_below = 5)
phrase_units <- segment(abstracts, level = "phrase", merge_below = 5)
```

```{r encode}
sentence_embeddings <- encode(sentence_units$text, model = qa_minilm)
clause_embeddings <- encode(clause_units$text, model = qa_minilm)
phrase_embeddings <- encode(phrase_units$text, model = qa_minilm)
```

```{r representatives}
evidence_sentence <- representatives(
  topic_model, sentence_units$text, embeddings = sentence_embeddings
)
evidence_clause <- representatives(
  topic_model, clause_units$text, embeddings = clause_embeddings
)
evidence_phrase <- representatives(
  topic_model, phrase_units$text, embeddings = phrase_embeddings
)
```

# What each level produces

Finer levels multiply the candidate pool and shrink the units — the
statistics below quantify the trade before you read a single statement.
`units` counts the candidate pool each level offers, `median_chars` the
typical unit length, and `median_statement_chars` the typical length of the
evidence actually selected across the thirty topics.

```{r overview}
data.frame(
  level = c("sentence", "clause", "phrase"),
  units = c(nrow(sentence_units), nrow(clause_units), nrow(phrase_units)),
  median_chars = c(
    median(nchar(sentence_units$text)),
    median(nchar(clause_units$text)),
    median(nchar(phrase_units$text))
  ),
  median_statement_chars = c(
    median(nchar(evidence_sentence$text)),
    median(nchar(evidence_clause$text)),
    median(nchar(evidence_phrase$text))
  )
)
```

# The statements, topic by topic

For every topic the three blocks below hold the three highest-margin
statements at each level, verbatim, with the distance to the topic centroid
in parentheses. Read them as answers to the same question — "show me the
unit that is unambiguously about this topic" — asked of three differently
cut candidate pools.

```{r panels, echo=FALSE, results="asis"}
show_level <- function(evidence, topic_id, level_name) {
  statements <- subset(evidence, topic == topic_id)
  cat(sprintf("**%s**\n\n", level_name))
  cat(sprintf(
    "> %s *(%.3f)*\n\n",
    statements$text,
    statements$distance
  ), sep = "")
}
invisible(lapply(
  seq_len(nrow(topic_model$topics)),
  function(topic_id) {
    topic_row <- subset(topic_model$topics, topic == topic_id)
    cat(sprintf("# Topic %d — %s\n\n", topic_id, topic_row$label))
    show_level(evidence_sentence, topic_id, "Sentence level")
    show_level(evidence_clause, topic_id, "Clause level")
    show_level(evidence_phrase, topic_id, "Phrase level")
  }
))
```

# Reading the comparison

Three things to look for. First, where sentence and clause agree, the
sentence had no internal structure to cut — the clause segmenter passed it
through whole, and the levels are interchangeable for that topic. Second,
where they disagree, the clause statement is usually the *core claim* of a
longer sentence whose margin improved once the subordinate material was
cut away. Third, phrase-level statements are the shortest but the most
fragile: a fragment can score a wide margin on vocabulary alone while no
longer being a claim anyone could check. That failure mode is why clause
is the default.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
