---
title: "Put Together: Blended vs Pure Sentence Topics, Matched Pair by Pair"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressMessages(devtools::load_all("..", quiet = TRUE))
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# What this page is

The blended vignette and the sentence vignette each show thirty topics —
but they are two *different* sets of thirty, so their topic numbers cannot
be compared across pages. This page puts them together properly: both
models are fitted here on the **identical** 99,765 sentences, from the
**identical** sentence encodings — the only difference between them is
whether each sentence's vector was blended with its abstract
(`sbert_blend()`, `alpha = 0.5`) or used as is. Every blended topic is
then matched to the pure-sentence topic it shares the most sentences
with, and the pairs are shown side by side: same label style, same
margin-ranked verbatim statements.

How to read a pair: the *shared* percentage says how much the two views
agree that these sentences belong together. High-share pairs are the same
topic seen twice — context changed nothing. Low-share pairs are where the
two designs genuinely disagree: the pure model grouped sentences that
*sound alike*, the blended model grouped sentences whose *papers are
about the same thing*.

# One corpus, two spaces

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
abstracts <- unique(trimws(usable$AB))
sentences <- sbert_segment(abstracts, level = "sentence", merge_below = 5)
nrow(sentences)
```

Both spaces are built from one pass of encodings, so the blend is the
only moving part:

```{r embeddings}
qa_minilm <- sbert_load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
sentence_embeddings <- sbert_encode(sentences$text, model = qa_minilm)
abstract_embeddings <- sbert_encode(abstracts, model = qa_minilm)
blended_embeddings <- sbert_blend(
  sentences,
  embeddings = sentence_embeddings,
  document_embeddings = abstract_embeddings,
  alpha = 0.5
)
```

# Two thirty-topic models

```{r models}
sentence_model <- sbert_topics(
  sentences$text,
  n_topics = 30,
  embeddings = sentence_embeddings,
  n_terms = 8,
  n_representatives = 3
)
blended_model <- sbert_topics(
  sentences$text,
  n_topics = 30,
  embeddings = blended_embeddings,
  n_terms = 8,
  n_representatives = 3
)
```

## Which space makes more coherent topics?

The single number to compare across the two tables is the mean NPMI: it
answers whether blending in document context produced topics whose top
words genuinely co-occur — the empirical verdict on the design.

```{r coherence}
sentence_coherence <- sbert_coherence(sentence_model, measure = "npmi")
blended_coherence <- sbert_coherence(blended_model, measure = "npmi")
mean(sentence_coherence$coherence)
mean(blended_coherence$coherence)
```

```{r evidence}
sentence_evidence <- sbert_representatives(
  sentence_model,
  sentences$text,
  embeddings = sentence_embeddings
)
blended_evidence <- sbert_representatives(
  blended_model,
  sentences$text,
  embeddings = blended_embeddings
)
```

# The pairing

Each blended topic is matched to the pure-sentence topic containing the
largest number of its sentences; `share` is the fraction of the blended
topic's sentences that the matched topic holds.

```{r align-helper, include=FALSE}
align_topics <- function(from_model, to_model) {
  overlap <- table(from_model$documents$topic, to_model$documents$topic)
  best <- apply(overlap, 1L, which.max)
  shared <- as.integer(overlap[cbind(seq_len(nrow(overlap)), best)])
  data.frame(
    blended_topic = as.integer(rownames(overlap)),
    sentence_topic = as.integer(colnames(overlap))[best],
    shared_sentences = shared,
    share = round(shared / as.numeric(rowSums(overlap)), 3),
    stringsAsFactors = FALSE
  )
}
```

```{r align}
pairing <- align_topics(blended_model, sentence_model)
pairing
```

# The pairs, side by side

```{r pair-helper, include=FALSE}
show_side <- function(model, evidence, topic_id, heading) {
  topic_row <- subset(model$topics, topic == topic_id)
  statements <- subset(evidence, topic == topic_id)
  cat(sprintf(
    "**%s — %s** (%d sentences)\n\n",
    heading,
    topic_row$label,
    topic_row$n_documents
  ))
  cat(sprintf(
    "> %s *(%.3f)*\n\n",
    statements$text,
    statements$distance
  ), sep = "")
}
```

```{r pairs, echo=FALSE, results="asis"}
invisible(lapply(
  seq_len(nrow(pairing)),
  function(pair_index) {
    pair <- subset(pairing, blended_topic == pair_index)
    cat(sprintf(
      "# Pair %d — %.0f%% shared\n\n",
      pair_index,
      100 * pair$share
    ))
    show_side(
      blended_model,
      blended_evidence,
      pair$blended_topic,
      "Blended (sentence + its abstract)"
    )
    show_side(
      sentence_model,
      sentence_evidence,
      pair$sentence_topic,
      "Pure sentence"
    )
  }
))
```

# What to conclude

Where a pair shares most of its sentences, context made no difference —
those topics are robust to the design choice, which is itself worth
knowing. Where the share is low, read the two statement sets against each
other: if the blended side reads like a research area and the pure side
like a rhetorical move ("we surveyed...", "results show..."), the blend
is doing what it was built to do — letting each sentence borrow the
subject of its paper. The mean NPMI above says which behavior dominates
on this corpus.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
