---
title: "Bibliographic Topic Modeling: The MCSE Corpus"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  dpi = 110,
  out.width = "100%"
)
pkgload::load_all("..", quiet = TRUE)
```

A working tutorial on a real bibliographic export: 16,453 computing-education
records with titles, abstracts, author keywords and metadata. Unlike the
packaged `topic-modeling` vignette, this one needs the MCSE file, so it lives
in `analysis/` rather than `vignettes/`.

Every step below is one verb with named arguments returning a data frame you
print. Nothing is assembled by hand at the call site.

## Loading

```{r load}
mcse_path <- Sys.getenv(
  "MCSE_RDS_PATH",
  unset = file.path(
    "/Users/mohammedsaqr/Documents/Documents X/MyData/AKA/STM",
    "MCSE_gold.RDS"
  )
)
mcse <- readRDS(mcse_path)

dim(mcse)
```

A Scopus/WoS export uses short field codes: `TI` title, `AB` abstract, `DE`
author keywords, `PY` year, `SO` source.

```{r fields}
head(subset(mcse, select = c(TI, PY, SO)), 3)
```

## Feeding the data frame in whole

Pass the frame and name the text column. Every other column rides along into
the fitted model, and rows whose abstract is missing, blank, or the literal
`[NO ABSTRACT AVAILABLE]` placeholder are dropped once — you never write that
filter yourself.

```{r encode, eval=FALSE}
fit <- topics(mcse, column = "AB", n_topics = 40)
```

Encoding 16,453 abstracts takes roughly twelve minutes on one core, so this
tutorial reuses a cached matrix. Encode once, keep the matrix, and record
*which model made it* — a matrix without its model id cannot be checked later,
and mixing revisions fails silently because cosines still compute.

```{r cached}
cache <- readRDS("../outputs/mcse_abstract_30_topics/mcse_abstract_embeddings.rds")

cache$model_id
cache$model_revision
dim(cache$abstract_embeddings)
```

The cache holds one row per *distinct* cleaned abstract — 15,308 of them, since
352 records duplicate another. The recorded assignments map each source row
back to its abstract:

```{r align}
assignments <- read.csv(
  "../outputs/mcse_abstract_30_topics/source_assignments.csv",
  stringsAsFactors = FALSE
)
modelled <- subset(assignments, included)

nrow(modelled)
```

```{r expand}
expand_rows <- function(embeddings, row_index) {
  embeddings[row_index, , drop = FALSE]
}
embeddings <- expand_rows(
  cache$abstract_embeddings,
  row_index = modelled$abstract_id
)
abstracts <- modelled$cleaned_abstract

dim(embeddings)
```

`expand_rows()` is a two-line local helper, not a call-site ritual: the
brackets live inside a function, which is the only place they belong.

## Choosing the topic count

```{r sweep, cache=TRUE}
sweep <- select_topics(
  abstracts,
  n_topics = c(10, 20, 30, 40, 50),
  embeddings = embeddings,
  measure = "npmi"
)

sweep
```

```{r sweep-plot}
plot(sweep)
```

On this corpus coherence climbs to 40 topics and then turns down, while
diversity and explained variance keep rising — as they always will with more
topics. Forty is where the evidence stops improving.

One caveat worth respecting: if any fit warns `did not converge in 100
iterations`, raise `iter_max` before trusting a close ranking.

```{r converge, eval=FALSE}
sweep <- select_topics(
  abstracts,
  n_topics = c(30, 40, 50),
  embeddings = embeddings,
  iter_max = 300
)
```

## The model

```{r fit}
fit <- fitted(sweep, n_topics = 40)
fit
```

```{r summary}
summary(fit)
```

## Reading the topics

Distinctive terms — words that separate this topic from its neighbours:

```{r terms}
head(terms(fit, n = 6), 18)
```

The words each topic uses most, which is a different question:

```{r terms-beta}
head(terms(fit, n = 6, sort_by = "beta"), 18)
```

In a single-discipline corpus the shared vocabulary swamps everything:
"students", "computer", "science", "course" appear in nearly every abstract, so
they describe no topic in particular. Remove them and the labels sharpen —
without refitting, because terms depend only on the assignments and the text.

```{r stopwords}
head(
  terms(
    fit,
    n = 6,
    stop_words = stop_words(
      add = c("students", "student", "computer", "science", "course",
              "courses", "computing", "education", "paper", "study")
    )
  ),
  18
)
```

BM25 weighting penalises cross-topic words more aggressively still:

```{r bm25}
head(terms(fit, n = 6, weighting = "bm25"), 12)
```

## Evidence

```{r representatives}
head(representatives(fit, n = 2), 10)
```

Margin ranking picks abstracts that belong to their own topic much more than to
the runner-up, so these are the clear cases rather than merely the central
ones.

## Quality per topic

```{r coherence}
head(coherence(fit, measure = "npmi"), 12)
```

```{r diversity}
topic_diversity(fit)
```

A low-coherence topic is a real finding, not a failure — it usually marks a
residual cluster the embedding could not separate.

## Structure among topics

```{r hierarchy, fig.height=7}
tree <- topic_hierarchy(fit)
plot(tree)
```

```{r reduce}
smaller <- reduce_topics(fit, n_topics = 20)
smaller
```

Merging down the tree keeps the 40-topic structure nested inside the 20, which
reads better in a report than an independent 20-topic fit.

## Metadata came along

Because the data frame went in whole, the year and source sit beside the topic
assignment, so prevalence over time needs no join:

```{r metadata, eval=FALSE}
fit_with_metadata <- topics(mcse, column = "AB", n_topics = 40)

head(
  subset(
    fit_with_metadata$documents,
    select = c(document_id, topic, label, PY, SO)
  ),
  8
)
```

Topic prevalence *over time* is the obvious next question, and it is the one
step this tutorial cannot write as a single verb. Counting documents per topic
per year currently means assembling a `table()` by hand at the call site —
exactly the ritual Rule #0 forbids. That is an API gap, not a user problem:
`TODO.md` lists dynamic topics (prevalence and per-period terms over a time
variable, holding assignments fixed) as planned work. Until that verb exists,
the honest thing is to note the gap rather than teach the workaround.

## What this corpus taught us

Three findings from testing retrieval and topic quality across this corpus and
two unrelated ones (a social-network-analysis corpus and a biomedical one):

- **Rank on abstracts, in one pass.** A title-first funnel — rank titles, keep
  the top few thousand, re-rank those on abstracts — never beat single-pass
  abstract scoring on any corpus tested. Whatever the title stage discards
  cannot be recovered, so stage-one recall caps the whole pipeline.
- **Weighting title, abstract and keyword scores together does not
  generalize.** Weights fitted on MCSE gave a small gain here, none on the
  second corpus, and a small gain on the third. The best weights differ per
  corpus. Abstract-only is the honest default.
- **Watch for topics that split on missing metadata.** When keyword vectors are
  mixed into the document vector, the best-scoring weightings on this corpus
  turned out to be separating records that *have* author keywords from those
  that do not — 26% of them. Coherence and diversity both reward that
  artifact, so neither metric catches it alone.

## Next

- `?terms` — every term setting, changeable without refitting.
- `?select_topics` — the sweep and its retained models.
- `blend()` with `segment()` — sentence-level topics that keep their parent
  abstract's context.
