---
title: "Sentence-Level CSE Topics — Words, c-TF-IDF, and Sentences"
subtitle: "Per-topic breakdown over 94,049 abstract sentences"
date: "`r format(Sys.Date())`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
report_data <- readRDS("../outputs/mcse_sentence_topics/bars_report_data.rds")
fig_uri <- function(kind, t) {
  knitr::image_uri(file.path("..", "tmp", "mcse_sentence_bars",
                             sprintf("%s_%02d.png", kind, t)))
}
```

Each sentence-level topic is described three ways, exactly as in the earlier
reviews. The **actual word count** (left bar) is the frequency view — what STM
calls the highest-probability words (β). The **c-TF-IDF** (right bar) is the
*distinctive* view (FREX analogue) — words common *within* the topic and rare
*across* topics. The **representative sentences** are the abstract sentences
nearest the cluster centre. Topics are ordered by size.

```{r cards, results='asis'}
for (topic in report_data) {
  cat(sprintf("\n## %s\n\n", topic$label))
  cat(sprintf("*%s sentences &middot; %.1f%% &middot; NPMI %.2f*\n\n",
              format(topic$n, big.mark = ","), 100 * topic$proportion, topic$npmi))
  cat(sprintf('<img src="%s" style="width:49%%;display:inline-block"> ',
              fig_uri("raw", topic$topic)))
  cat(sprintf('<img src="%s" style="width:49%%;display:inline-block">\n\n',
              fig_uri("tfidf", topic$topic)))
  cat("**Representative sentences**\n\n")
  cat(paste(sprintf("- %s", topic$reps), collapse = "\n"), "\n\n")
}
```
