---
title: "MCSE Abstract Topics"
subtitle: "Review of a forced 30-topic semantic clustering of canonical abstracts"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 120
)

artifact_relative_path <- file.path(
  "outputs",
  "mcse_abstract_30_topics",
  "mcse_abstract_30_topics.rds"
)
artifact_candidates <- c(
  artifact_relative_path,
  file.path("..", artifact_relative_path)
)
artifact_exists <- file.exists(artifact_candidates)
if (!any(artifact_exists)) {
  stop(
    paste0(
      "The completed MCSE topic artifact is not available. Expected ",
      artifact_relative_path,
      ". Run analysis/mcse_abstract_30_topics.R before knitting this review."
    ),
    call. = FALSE
  )
}

result_path <- artifact_candidates[which(artifact_exists)[[1L]]]
project_root <- normalizePath(
  file.path(dirname(result_path), "..", ".."),
  mustWork = TRUE
)
source(file.path(
  project_root,
  "analysis",
  "sentence_level_topic_functions.R"
))
source(file.path(project_root, "R", "topics.R"))
result <- readRDS(result_path)
required_components <- c(
  "source_assignments", "canonical_abstracts", "topic_summary",
  "representatives", "boundary_abstracts", "frequent_ngrams",
  "review_queue", "restart_diagnostics", "data_audit", "model", "method"
)
stopifnot(
  is.list(result),
  all(required_components %in% names(result)),
  is.data.frame(result$source_assignments),
  is.data.frame(result$canonical_abstracts),
  is.data.frame(result$topic_summary),
  is.data.frame(result$representatives),
  is.data.frame(result$boundary_abstracts),
  is.data.frame(result$frequent_ngrams),
  is.data.frame(result$restart_diagnostics),
  is.data.frame(result$data_audit),
  is.data.frame(result$method)
)

assignments <- result$source_assignments
abstracts <- result$canonical_abstracts
topics <- result$topic_summary
representatives <- result$representatives
boundaries <- result$boundary_abstracts
ngrams <- result$frequent_ngrams
restarts <- result$restart_diagnostics

stopifnot(
  nrow(assignments) == 16453L,
  sum(assignments$included) == 15660L,
  nrow(abstracts) == 15308L,
  nrow(topics) == 30L,
  identical(sort(topics$topic), seq_len(30L)),
  sum(topics$n_source_records) == 15660L,
  sum(topics$n_distinct_abstracts) == 15308L,
  nrow(representatives) == 120L,
  nrow(boundaries) == 60L,
  nrow(ngrams) == 900L,
  sum(restarts$selected) == 1L,
  all(is.finite(topics$median_cosine_distance)),
  all(is.finite(boundaries$assignment_margin))
)

audit_value <- setNames(result$data_audit$value, result$data_audit$metric)

format_count <- function(value) {
  stopifnot(is.numeric(value), !anyNA(value))
  format(value, big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_percent <- function(value, digits = 1L) {
  stopifnot(
    is.numeric(value),
    !anyNA(value),
    is.numeric(digits),
    length(digits) == 1L,
    digits == as.integer(digits)
  )
  sprintf(paste0("%.", digits, "f%%"), 100 * value)
}

html_escape <- function(value) {
  stopifnot(is.character(value), !anyNA(value))
  value |>
    gsub("&", "&amp;", x = _, fixed = TRUE) |>
    gsub("<", "&lt;", x = _, fixed = TRUE) |>
    gsub(">", "&gt;", x = _, fixed = TRUE) |>
    gsub('"', "&quot;", x = _, fixed = TRUE) |>
    gsub("'", "&#39;", x = _, fixed = TRUE)
}

compact_text <- function(value, maximum = 150L) {
  stopifnot(
    is.character(value),
    !anyNA(value),
    is.numeric(maximum),
    length(maximum) == 1L,
    maximum >= 4L
  )
  value <- trimws(gsub("[[:space:]]+", " ", value))
  ifelse(
    nchar(value, type = "chars") <= maximum,
    value,
    paste0(substr(value, 1L, maximum - 1L), "…")
  )
}

stopifnot(
  identical(format_count(1234), "1,234"),
  identical(format_percent(0.125), "12.5%"),
  identical(html_escape("<&"), "&lt;&amp;"),
  identical(compact_text("abcdef", 5L), "abcd…")
)
```

<style>
body { color: #25364a; font-size: 15px; }
h1, h2, h3, h4 { color: #153e5c; }
.kpi-grid { display: grid; grid-template-columns: repeat(5, minmax(135px, 1fr)); gap: 12px; margin: 18px 0 24px; }
.kpi { background: #f3f7fa; border-top: 4px solid #247ba0; border-radius: 6px; padding: 13px 15px; }
.kpi-value { color: #153e5c; font-size: 25px; font-weight: 700; }
.kpi-label { color: #526579; font-size: 11px; letter-spacing: .04em; text-transform: uppercase; }
.caveat { background: #fff7dd; border-left: 6px solid #d99000; border-radius: 5px; padding: 15px 18px; margin: 18px 0 24px; }
.method-note { background: #edf8f5; border-left: 6px solid #238b78; border-radius: 5px; padding: 15px 18px; margin: 18px 0 24px; }
.topic-block { border: 1px solid #d6e0e8; border-radius: 7px; padding: 14px 18px; margin: 16px 0; background: #fff; }
.topic-block h3 { margin-top: 0; }
.topic-meta { color: #5d6d7e; margin-bottom: 12px; }
.canonical-label { background: #f4f7f9; border-left: 4px solid #247ba0; padding: 10px 13px; margin: 10px 0; }
details { border-top: 1px solid #e3e8ed; padding: 9px 0; }
summary { color: #153e5c; cursor: pointer; font-weight: 600; }
.paper-meta { color: #5d6d7e; font-size: 13px; margin: 7px 0; }
.abstract { white-space: normal; line-height: 1.45; }
.secondary { background: #f7f7f7; border-left: 5px solid #6b7280; padding: 13px 16px; margin: 14px 0; }
table { font-size: 12px; }
th { background: #247ba0 !important; color: #fff !important; }
@media (max-width: 900px) { .kpi-grid { grid-template-columns: repeat(2, minmax(130px, 1fr)); } }
</style>

## Corpus audit

```{r audit-kpis, results='asis'}
kpi_values <- c(
  format_count(audit_value[["source_rows"]]),
  format_count(audit_value[["modeled_source_rows"]]),
  format_count(audit_value[["distinct_clean_abstracts"]]),
  format_count(audit_value[["missing_abstracts"]]),
  format_count(audit_value[["placeholder_abstracts"]])
)
kpi_labels <- c(
  "Source records", "Modeled records", "Distinct cleaned abstracts",
  "Missing abstracts", "Placeholder abstracts"
)
kpi_items <- paste0(
  '<div class="kpi"><div class="kpi-value">',
  kpi_values,
  '</div><div class="kpi-label">',
  kpi_labels,
  '</div></div>'
)
cat('<div class="kpi-grid">', paste(kpi_items, collapse = ""), '</div>')
```

The analysis excludes missing abstracts and the literal placeholder
`[NO ABSTRACT AVAILABLE]`. It fits each distinct cleaned abstract once, then
maps the assignment back to every source record. Long abstracts are represented
by sentence-aware chunks combined into one normalized abstract embedding.

```{r audit-table}
audit_display <- result$data_audit
audit_display$value <- format_count(audit_display$value)
knitr::kable(
  audit_display,
  col.names = c("Audit metric", "Value"),
  row.names = FALSE
)
```

<div class="caveat">
<strong>Forced topic-count caveat.</strong> Thirty topics were fixed by the
analysis specification. They were not inferred as a uniquely optimal number
from the corpus. Topic IDs are arbitrary and run-specific. Each automatic label
is the exact lead sentence of the centroid-nearest corpus abstract—not a
validated taxonomy or a generated summary. No c-TF-IDF or keyword score was
used for assignments or labels.
</div>

## Topic size and label overview

```{r topic-size-plot}
topic_order <- order(topics$n_source_records)
old_par <- graphics::par(mar = c(5, 5, 3, 1))
graphics::barplot(
  topics$n_source_records[topic_order],
  names.arg = paste0("T", topics$topic[topic_order]),
  horiz = TRUE,
  las = 1,
  col = "#247ba0",
  border = NA,
  xlab = "Source records",
  main = "Source-record prevalence across 30 forced topics"
)
graphics::par(old_par)
```

```{r topic-overview-table}
topic_display <- topics[
  , c(
    "topic", "canonical_lead_sentence", "canonical_paper_title",
    "n_source_records", "source_record_share", "n_distinct_abstracts",
    "median_cosine_distance", "p90_cosine_distance", "year_min", "year_max"
  )
]
names(topic_display) <- c(
  "Topic", "Exact corpus label", "Canonical paper", "Source records",
  "Share", "Distinct abstracts", "Median distance", "P90 distance",
  "First year", "Last year"
)
topic_display$`Exact corpus label` <- compact_text(
  topic_display$`Exact corpus label`,
  180L
)
topic_display$`Canonical paper` <- compact_text(
  topic_display$`Canonical paper`,
  100L
)
topic_display$Share <- format_percent(topic_display$Share)
topic_display$`Median distance` <- sprintf("%.3f", topic_display$`Median distance`)
topic_display$`P90 distance` <- sprintf("%.3f", topic_display$`P90 distance`)
knitr::kable(topic_display, row.names = FALSE)
```

Cosine distance is not a probability or confidence score. Smaller values mean
that an abstract is closer to its assigned topic centroid. Assignment margin is
the selected-centroid similarity minus the second-best similarity; smaller
margins identify less distinct assignments.

## Representative papers and canonical abstracts

The four representatives for each topic were selected from the central quarter
of that topic. Rank 1 is the centroid-nearest canonical abstract. Ranks 2–4 add
embedding-space diversity through maximal marginal relevance. The labels and
representatives therefore remain semantic, full-abstract evidence.

```{r representative-abstracts, results='asis'}
representative_topic_html <- vapply(
  seq_len(nrow(topics)),
  function(topic_row) {
    topic_id <- topics$topic[[topic_row]]
    topic_representatives <- representatives[
      representatives$topic == topic_id,
      ,
      drop = FALSE
    ]
    topic_representatives <- topic_representatives[
      order(topic_representatives$rank),
      ,
      drop = FALSE
    ]
    paper_html <- vapply(
      seq_len(nrow(topic_representatives)),
      function(representative_row) {
        paper <- topic_representatives[representative_row, , drop = FALSE]
        doi_text <- if (
          is.na(paper$doi[[1L]]) || !nzchar(trimws(paper$doi[[1L]]))
        ) {
          "DOI unavailable"
        } else {
          paste0("DOI: ", html_escape(paper$doi[[1L]]))
        }
        sprintf(
          paste0(
            '<details%s><summary>Rank %d%s · %s (%s)</summary>',
            '<div class="paper-meta">%s · %s · %s · %s · distance %.3f · margin %.4f</div>',
            '<p class="abstract">%s</p></details>'
          ),
          if (paper$rank[[1L]] == 1L) " open" else "",
          paper$rank[[1L]],
          if (paper$rank[[1L]] == 1L) " — canonical" else "",
          html_escape(paper$title[[1L]]),
          html_escape(as.character(paper$year[[1L]])),
          html_escape(paper$authors[[1L]]),
          html_escape(ifelse(is.na(paper$source[[1L]]), "Source unavailable", paper$source[[1L]])),
          doi_text,
          html_escape(paper$source_record_id[[1L]]),
          paper$cosine_distance_to_centroid[[1L]],
          paper$assignment_margin[[1L]],
          html_escape(paper$cleaned_abstract[[1L]])
        )
      },
      character(1)
    )
    sprintf(
      paste0(
        '<div class="topic-block"><h3>Topic %d</h3>',
        '<div class="topic-meta">%s source records · %s distinct abstracts · %s</div>',
        '<div class="canonical-label"><strong>Exact corpus label:</strong> %s</div>',
        '%s</div>'
      ),
      topic_id,
      format_count(topics$n_source_records[[topic_row]]),
      format_count(topics$n_distinct_abstracts[[topic_row]]),
      format_percent(topics$source_record_share[[topic_row]]),
      html_escape(topics$canonical_lead_sentence[[topic_row]]),
      paste(paper_html, collapse = "\n")
    )
  },
  character(1)
)
cat(paste(representative_topic_html, collapse = "\n"))
```

## Boundary cases

These are the two lowest-margin canonical abstracts assigned to each topic.
They are a qualitative stress test of the 30-way partition. A low margin means
that the first- and second-best centroids were similar; it does not estimate an
error probability.

```{r boundary-table}
boundary_display <- boundaries[
  , c(
    "topic", "rank", "title", "year", "source_record_id",
    "cosine_distance_to_centroid", "assignment_margin", "cleaned_abstract"
  )
]
names(boundary_display) <- c(
  "Topic", "Boundary rank", "Paper title", "Year", "Record ID",
  "Cosine distance", "Assignment margin", "Abstract preview"
)
boundary_display$`Paper title` <- compact_text(boundary_display$`Paper title`, 90L)
boundary_display$`Abstract preview` <- compact_text(
  boundary_display$`Abstract preview`,
  180L
)
boundary_display$`Cosine distance` <- sprintf(
  "%.3f",
  boundary_display$`Cosine distance`
)
boundary_display$`Assignment margin` <- sprintf(
  "%.4f",
  boundary_display$`Assignment margin`
)
knitr::kable(boundary_display, row.names = FALSE)
```

The complete review queue also includes excluded source records and every
canonical abstract in the lowest 10% of assignment margins. It remains in the
saved RDS and exported CSV rather than being printed in full here.

## Restart diagnostics

Fifty seeded spherical k-means restarts were evaluated. The selected restart is
the size-balanced candidate with the highest normalized size entropy among
solutions within 0.5% of the best cosine objective and with a minimum topic size
of at least 30. Non-converged restarts are never eligible.

Adjusted Rand Index (ARI) compares every restart with the selected partition
without depending on arbitrary topic numbers. An ARI of 1 means the partitions
are identical; values near 0 indicate chance-level agreement.

```{r restart-plots}
selected_restart <- restarts$selected
old_par <- graphics::par(mfrow = c(1, 2), mar = c(5, 5, 3, 1))
graphics::plot(
  restarts$seed,
  restarts$objective,
  pch = ifelse(selected_restart, 19, 1),
  col = ifelse(selected_restart, "#d94841", "#7b8794"),
  xlab = "Restart seed",
  ylab = "Total assigned cosine similarity",
  main = "Objective across restarts"
)
graphics::points(
  restarts$seed[selected_restart],
  restarts$objective[selected_restart],
  pch = 19,
  col = "#d94841",
  cex = 1.3
)
graphics::plot(
  restarts$objective,
  restarts$normalized_size_entropy,
  pch = ifelse(selected_restart, 19, 1),
  col = ifelse(selected_restart, "#d94841", "#247ba0"),
  xlab = "Total assigned cosine similarity",
  ylab = "Normalized topic-size entropy",
  main = "Fit versus size balance"
)
graphics::points(
  restarts$objective[selected_restart],
  restarts$normalized_size_entropy[selected_restart],
  pch = 19,
  col = "#d94841",
  cex = 1.3
)
graphics::par(old_par)
```

```{r restart-table}
restart_order <- order(
  -restarts$selected,
  -restarts$objective,
  restarts$seed
)
restart_display <- utils::head(restarts[restart_order, , drop = FALSE], 12L)
restart_display$objective <- sprintf("%.4f", restart_display$objective)
restart_display$normalized_size_entropy <- sprintf(
  "%.4f",
  restart_display$normalized_size_entropy
)
knitr::kable(restart_display, row.names = FALSE)
```

## Raw n-grams — secondary audit only

<div class="secondary">
<strong>Not part of the topic model.</strong> These raw unigram, bigram, and
trigram counts were calculated after assignment. They were not used to create
embeddings, fit centroids, assign abstracts, select representatives, or produce
automatic labels. They are shown only as a lexical audit of each semantic
cluster.
</div>

### Most common words by topic

Each panel shows the eight most frequent remaining unigrams in the distinct
abstracts assigned to that topic. Standard English stop words and ubiquitous
corpus terms—including *computer*, *students*, *coding*, *programming*,
*computing*, and *education*—are excluded so the panels reveal topic-specific
language. Research boilerplate such as *paper*, *study*, *course*, *results*,
*used*, and *using* is excluded for the same reason.

Bar lengths remain raw word-occurrence counts, not c-TF-IDF weights or document
frequencies. Longer abstracts can therefore contribute more occurrences. The
additional exclusions affect only this descriptive plot; embeddings and topic
assignments are unchanged.

```{r common-word-bars, fig.width=14, fig.height=28}
common_word_stopwords <- unique(c(
  stop_words(),
  mcse_common_word_stopwords()
))
common_word_tables <- lapply(
  seq_len(30L),
  function(topic_id) {
    frequent_sentence_ngrams(
      abstracts$cleaned_abstract[abstracts$topic == topic_id],
      n = 1L,
      top_n = 8L,
      stopwords = common_word_stopwords
    )
  }
)
common_word_data <- do.call(
  rbind,
  lapply(
    seq_along(common_word_tables),
    function(topic_id) {
      data.frame(
        topic = rep.int(topic_id, nrow(common_word_tables[[topic_id]])),
        common_word_tables[[topic_id]],
        stringsAsFactors = FALSE
      )
    }
  )
)
common_word_data <- common_word_data[
  order(common_word_data$topic, common_word_data$rank),
  ,
  drop = FALSE
]
stopifnot(
  nrow(common_word_data) == 240L,
  identical(sort(unique(common_word_data$topic)), seq_len(30L)),
  identical(
    as.integer(table(common_word_data$topic)),
    rep.int(8L, 30L)
  ),
  all(common_word_data$frequency > 0L),
  !any(common_word_data$ngram %in% mcse_common_word_stopwords()),
  all(vapply(
    split(common_word_data$rank, common_word_data$topic),
    identical,
    logical(1),
    seq_len(8L)
  ))
)

old_par <- graphics::par(
  mfrow = c(6, 5),
  mar = c(2.4, 7.2, 2.5, 0.7),
  oma = c(3.5, 0.5, 2, 0.5),
  xaxs = "i"
)
plot_positions <- lapply(
  seq_len(30L),
  function(topic_id) {
    topic_words <- common_word_data[
      common_word_data$topic == topic_id,
      ,
      drop = FALSE
    ]
    graphics::barplot(
      rev(topic_words$frequency),
      names.arg = rev(topic_words$ngram),
      horiz = TRUE,
      las = 1,
      col = "#247ba0",
      border = NA,
      cex.names = 0.78,
      cex.axis = 0.70,
      main = sprintf("Topic %02d", topic_id),
      xlab = ""
    )
  }
)
graphics::mtext(
  "Raw unigram occurrences after generic corpus-word removal",
  side = 1,
  outer = TRUE,
  line = 1.2,
  cex = 1.05
)
graphics::par(old_par)
stopifnot(
  length(plot_positions) == 30L,
  all(vapply(plot_positions, length, integer(1)) == 8L)
)
```

```{r ngram-sections, results='asis'}
ngram_topic_html <- vapply(
  seq_len(nrow(topics)),
  function(topic_row) {
    topic_id <- topics$topic[[topic_row]]
    topic_ngrams <- ngrams[
      ngrams$topic == topic_id & ngrams$rank <= 5L,
      c("n", "rank", "ngram", "frequency"),
      drop = FALSE
    ]
    names(topic_ngrams) <- c("N", "Rank", "Raw n-gram", "Frequency")
    table_html <- paste(
      knitr::kable(topic_ngrams, format = "html", row.names = FALSE),
      collapse = "\n"
    )
    sprintf(
      '<details><summary>Topic %d raw n-grams</summary>%s</details>',
      topic_id,
      table_html
    )
  },
  character(1)
)
cat(paste(ngram_topic_html, collapse = "\n"))
```

## Method record

```{r method-table}
knitr::kable(
  result$method,
  col.names = c("Method item", "Value"),
  row.names = FALSE
)
```

<div class="method-note">
This report is an interpretation aid, not evidence that 30 is the natural or
only defensible topic count. Review the canonical abstracts, diverse
representatives, and low-margin boundary papers together before assigning
short human-readable topic names or using the topics in substantive claims.
</div>
