---
title: "Sentence-level topics in translated feedback"
subtitle: "Six spherical sentence-embedding clusters defined by complete corpus sentences"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5,
  dpi = 120
)
result_path <- file.path(
  "outputs",
  "feedback_translation_topics",
  "sentence_level_topics",
  "sentence_level_topic_model.rds"
)
stopifnot(file.exists(result_path))
result <- readRDS(result_path)
topic_summary <- result$topic_summary
representatives <- result$representative_sentences
assignments <- result$sentence_assignments
stopifnot(
  nrow(assignments) == 8987L,
  sum(assignments$included) == 8976L,
  nrow(topic_summary) == 6L,
  nrow(representatives) == 30L,
  nrow(result$boundary_sentences) == 18L
)

format_percent <- function(value, digits = 1L) {
  stopifnot(is.numeric(value), is.numeric(digits), length(digits) == 1L)
  sprintf(paste0("%.", digits, "f%%"), 100 * value)
}
```

<style>
body { font-size: 16px; color: #1f2937; }
.verdict { padding: 18px 22px; border-left: 6px solid #0f766e; background: #f0fdfa; margin: 18px 0 26px; }
.cards { display: flex; flex-wrap: wrap; gap: 14px; margin: 18px 0 26px; }
.card { flex: 1 1 180px; padding: 15px 18px; border: 1px solid #d1d5db; border-radius: 8px; background: #f8fafc; }
.topic-card { padding: 18px 22px; border: 1px solid #cbd5e1; border-radius: 9px; margin: 16px 0; background: #ffffff; }
.topic-card h3 { color: #0f4c5c; margin-top: 0; }
.number { font-size: 28px; font-weight: 700; color: #0f4c5c; }
.label { color: #475569; }
.warning { padding: 12px 16px; background: #fffbeb; border-left: 5px solid #d97706; }
.prototype { padding: 10px 14px; background: #f1f5f9; font-style: italic; }
table { font-size: 14px; }
</style>

## Result

<div class="verdict">
The revised model is sentence-level end to end. It clusters complete MiniLM
sentence embeddings with 50 cosine-aware spherical k-means restarts. Every
automatic topic label is an exact sentence from the corpus, and each topic is
shown through diverse complete representative sentences. Ordinary word and
phrase frequencies appear only in a separate audit appendix.
</div>

```{r cards, results='asis'}
cat(sprintf(
  paste0(
    '<div class="cards">',
    '<div class="card"><div class="number">8,976</div><div class="label">modeled rows</div></div>',
    '<div class="card"><div class="number">8,144</div><div class="label">distinct sentences</div></div>',
    '<div class="card"><div class="number">6</div><div class="label">sentence topics</div></div>',
    '<div class="card"><div class="number">%s</div><div class="label">largest topic</div></div>',
    '</div>'
  ),
  format_percent(max(topic_summary$proportion))
))
```

<div class="warning">
The concise intent titles are provisional analyst interpretations. The exact
canonical corpus sentences are the automatic labels. Sixty-three distinct
language-review sentences were held out while fitting centroids and assigned
provisionally afterward; ambiguous boundary cases remain in the review queue.
</div>

## Six sentence-defined topics

```{r topic-cards, results='asis'}
topic_card_html <- vapply(
  seq_len(nrow(topic_summary)),
  function(topic_row) {
    topic_id <- topic_summary$topic[[topic_row]]
    topic_representatives <- representatives[
      representatives$topic == topic_id,
      ,
      drop = FALSE
    ]
    representative_items <- paste0(
      "<li>",
      htmltools::htmlEscape(topic_representatives$representative_sentence),
      "</li>",
      collapse = ""
    )
    sprintf(
      paste0(
        '<div class="topic-card">',
        '<h3>Topic %d — %s</h3>',
        '<p><strong>%s rows (%s)</strong> · %s distinct sentences</p>',
        '<div class="prototype">Canonical sentence: %s</div>',
        '<p><strong>Diverse full-sentence representatives</strong></p><ol>%s</ol>',
        '</div>'
      ),
      topic_id,
      htmltools::htmlEscape(topic_summary$intent_title[[topic_row]]),
      format(topic_summary$n_rows[[topic_row]], big.mark = ","),
      format_percent(topic_summary$proportion[[topic_row]]),
      format(topic_summary$n_unique_sentences[[topic_row]], big.mark = ","),
      htmltools::htmlEscape(topic_summary$canonical_sentence[[topic_row]]),
      representative_items
    )
  },
  character(1)
)
cat(paste(topic_card_html, collapse = "\n"))
```

## Topic sizes and cohesion

```{r size-plot}
topic_order <- order(topic_summary$n_rows)
barplot(
  topic_summary$n_rows[topic_order],
  names.arg = paste0(
    "T",
    topic_summary$topic[topic_order],
    ": ",
    topic_summary$intent_title[topic_order]
  ),
  horiz = TRUE,
  las = 1,
  col = "#0f766e",
  border = NA,
  xlab = "Number of rows",
  main = "Balanced six-topic sentence solution"
)
```

```{r summary-table}
summary_display <- topic_summary[
  , c(
    "topic", "intent_title", "n_rows", "proportion",
    "n_unique_sentences", "mean_cosine_distance", "p90_cosine_distance",
    "assigned_rows", "ambiguous_rows", "translation_review_rows",
    "canonical_sentence"
  )
]
summary_display$proportion <- format_percent(summary_display$proportion)
summary_display$mean_cosine_distance <- sprintf(
  "%.3f",
  summary_display$mean_cosine_distance
)
summary_display$p90_cosine_distance <- sprintf(
  "%.3f",
  summary_display$p90_cosine_distance
)
knitr::kable(summary_display, row.names = FALSE)
```

Cosine distance is not a probability. Smaller values indicate proximity to the
topic centroid. Assignment margin is the selected-centroid similarity minus the
second-best similarity; smaller margins indicate less certain boundaries.

## Boundary sentences

These are the three lowest-margin eligible sentences per topic. They expose
where the six-topic constraint is least convincing.

```{r boundary-table}
boundary_display <- result$boundary_sentences[
  , c(
    "topic", "intent_title", "rank", "boundary_sentence",
    "cosine_distance_to_centroid", "assignment_margin"
  )
]
boundary_display$cosine_distance_to_centroid <- sprintf(
  "%.3f",
  boundary_display$cosine_distance_to_centroid
)
boundary_display$assignment_margin <- sprintf(
  "%.4f",
  boundary_display$assignment_margin
)
knitr::kable(boundary_display, row.names = FALSE)
```

## Frequent language patterns — secondary only

The following unigrams, bigrams, and trigrams are ordinary within-topic counts.
They were not used for sentence assignment, centroid fitting, automatic labels,
or intent titles.

```{r ngram-table}
knitr::kable(result$frequent_ngrams, row.names = FALSE)
```

## Review queue

The queue contains 892 lowest-margin assignments, 63 language-review rows, and
11 excluded blanks. The first 250 rows are displayed; the complete queue is in
the exported CSV.

```{r review-table}
review_display <- utils::head(
  result$review_queue[
    , c(
      "row_id", "translation", "topic", "intent_title",
      "assignment_margin", "assignment_status"
    )
  ],
  250L
)
review_display$assignment_margin <- sprintf(
  "%.4f",
  review_display$assignment_margin
)
knitr::kable(review_display, row.names = FALSE)
```

## Restart diagnostics

The selected restart maximized total assigned cosine similarity across the
8,081 language-screened distinct sentences.

```{r restart-plot}
diagnostics <- result$restart_diagnostics
plot(
  diagnostics$seed,
  diagnostics$objective,
  type = "h",
  lwd = 3,
  col = ifelse(diagnostics$selected, "#dc2626", "#94a3b8"),
  xlab = "Restart seed",
  ylab = "Total cosine objective",
  main = "Fifty spherical k-means restarts"
)
points(
  diagnostics$seed[diagnostics$selected],
  diagnostics$objective[diagnostics$selected],
  pch = 19,
  col = "#dc2626"
)
```

## Method

```{r method-table}
knitr::kable(result$method, row.names = FALSE)
```
