---
title: "Six-topic SBERT vs R text/BERTopic"
subtitle: "Translated feedback: 8,976 modeled rows"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5,
  dpi = 120
)
comparison_path <- file.path(
  "outputs",
  "feedback_translation_topics",
  "text_package_comparison",
  "text_topic_comparison.rds"
)
stopifnot(file.exists(comparison_path))
comparison <- readRDS(comparison_path)
assignments <- comparison$comparison_assignments
quality <- comparison$quality_metrics
agreement <- comparison$agreement_metrics
topic_summaries <- comparison$topic_summaries
reduced_info <- comparison$reduced_topic_info
stopifnot(
  nrow(assignments) == 8976L,
  identical(sort(unique(assignments$sbert_topic)), 1:6),
  identical(sort(unique(assignments$text_reduced_topic)), c(-1L, 0:5))
)

format_percent <- function(value, digits = 1L) {
  stopifnot(is.numeric(value), length(value) >= 1L, is.numeric(digits))
  sprintf(paste0("%.", digits, "f%%"), 100 * value)
}

format_metric <- function(value, digits = 3L) {
  stopifnot(is.numeric(value), length(value) >= 1L, is.numeric(digits))
  sprintf(paste0("%.", digits, "f"), value)
}

all_rows_agreement <- subset(
  agreement,
  subset == "all_rows" & method_a == "sbert_6" &
    method_b == "text_reduced_6"
)
native_quality <- subset(quality, method == "text_native")
reduced_quality <- subset(quality, method == "text_reduced_6")
```

<style>
body { font-size: 16px; color: #1f2937; }
.verdict { padding: 18px 22px; border-left: 6px solid #b91c1c; background: #fef2f2; margin: 18px 0 28px; }
.cards { display: flex; flex-wrap: wrap; gap: 14px; margin: 18px 0 26px; }
.card { flex: 1 1 180px; padding: 15px 18px; border: 1px solid #d1d5db; border-radius: 8px; background: #f8fafc; }
.number { font-size: 28px; font-weight: 700; color: #0f4c5c; }
.label { color: #475569; }
.note { padding: 12px 16px; background: #fffbeb; border-left: 5px solid #d97706; }
table { font-size: 14px; }
</style>

## Verdict

<div class="verdict">
<strong>Your concern is justified.</strong> Replacing deterministic six-cluster
SBERT with <code>text::textTopics()</code> does not yield a clearer six-topic
solution. Native <code>text</code>/BERTopic fragments this repetitive corpus into
405 non-outlier topics and rejects 1,602 rows as outliers. Reducing that model to
six substantive topics then puts 89.1% of its modeled rows into one catch-all
topic. The two six-topic partitions have weak agreement (ARI 0.139).
</div>

```{r cards, results='asis'}
cat(sprintf(
  paste0(
    '<div class="cards">',
    '<div class="card"><div class="number">6</div><div class="label">SBERT forced topics</div></div>',
    '<div class="card"><div class="number">%d</div><div class="label">native text topics</div></div>',
    '<div class="card"><div class="number">%s</div><div class="label">native text outliers</div></div>',
    '<div class="card"><div class="number">%s</div><div class="label">largest reduced-text topic</div></div>',
    '</div>'
  ),
  native_quality$topic_count,
  format_percent(native_quality$outlier_share),
  format_percent(reduced_quality$largest_topic_share)
))
```

## What was compared

Both methods used the same 8,976 translated rows and the same pinned
`all-MiniLM-L6-v2` Sentence-BERT model revision. A 32-row cross-runtime audit
found a maximum absolute embedding difference of
`r format(comparison$method_details$value[comparison$method_details$item == "embedding_parity_max_abs_difference"], scientific = TRUE)`;
NumPy `allclose()` passed. Consequently, the differences below come from topic
construction, not materially different sentence vectors.

```{r methods-table}
method_table <- data.frame(
  Method = c("sbert_6", "text_native", "text_reduced_6"),
  Topic_construction = c(
    "Deterministic Lloyd k-means, exactly 6 clusters",
    "UMAP + HDBSCAN + BERTopic c-TF-IDF/MMR",
    "Native BERTopic topics merged post hoc to 6 substantive topics"
  ),
  Topics = quality$topic_count,
  Outliers = quality$outlier_rows,
  Outlier_share = format_percent(quality$outlier_share),
  Largest_topic_share = format_percent(quality$largest_topic_share),
  stringsAsFactors = FALSE
)
knitr::kable(method_table, row.names = FALSE)
```

<div class="note">
BERTopic preserves its outlier group during reduction. Requesting seven total
groups was therefore required to obtain six non-outlier topics plus topic −1.
</div>

## Agreement between outcomes

Adjusted Rand Index (ARI) is 1 for identical partitions up to label renaming and
near 0 when agreement is close to chance. Normalized mutual information (NMI)
ranges from 0 to 1. The results show that the two approaches impose substantially
different structures.

```{r agreement-table}
agreement_display <- agreement[
  agreement$subset %in% c("all_rows", "native_non_outliers") &
    agreement$method_a == "sbert_6" &
    agreement$method_b %in% c("text_native", "text_reduced_6"),
  c(
    "subset", "method_a", "method_b", "n_rows", "adjusted_rand",
    "normalized_mutual_information", "normalized_variation_information"
  )
]
agreement_display$adjusted_rand <- format_metric(
  agreement_display$adjusted_rand
)
agreement_display$normalized_mutual_information <- format_metric(
  agreement_display$normalized_mutual_information
)
agreement_display$normalized_variation_information <- format_metric(
  agreement_display$normalized_variation_information
)
knitr::kable(agreement_display, row.names = FALSE)
```

## Six-topic assignment overlap

Rows show the original six SBERT topics; columns show reduced `text` topics.
Column −1 contains native BERTopic outliers. The large column 0 absorbs most
SBERT topics 1–3, demonstrating why the reduced output is not a useful six-way
taxonomy.

```{r cross-tab}
overlap <- addmargins(table(
  SBERT = assignments$sbert_topic,
  text_reduced = assignments$text_reduced_topic
))
knitr::kable(as.data.frame.matrix(overlap), row.names = TRUE)
```

## Topic balance

```{r balance-plot}
balance <- topic_summaries[
  topic_summaries$method %in% c("sbert_6", "text_reduced_6"),
  c("method", "topic", "n_rows", "proportion_of_modeled")
]
balance$label <- paste(balance$method, balance$topic, sep = " / topic ")
balance <- balance[order(balance$n_rows), , drop = FALSE]
barplot(
  balance$n_rows,
  names.arg = balance$label,
  horiz = TRUE,
  las = 1,
  col = ifelse(balance$method == "sbert_6", "#0f766e", "#c2410c"),
  border = NA,
  xlab = "Modeled rows (outliers excluded)",
  main = "Topic-size imbalance"
)
legend(
  "bottomright",
  legend = c("sbert_6", "text_reduced_6"),
  fill = c("#0f766e", "#c2410c"),
  bty = "n"
)
```

## Common embedding-space quality

All distances and silhouettes below were calculated in the same original
384-dimensional normalized MiniLM embedding space. Native `text` forms tighter
small clusters, but it does so by creating 405 topics and excluding 17.8% of
rows. After reduction to six, its cohesion is worse than the original six-cluster
solution and its size balance collapses.

```{r quality-table}
quality_display <- quality[
  , c(
    "method", "topic_count", "modeled_rows", "outlier_rows",
    "largest_topic_share", "normalized_size_entropy",
    "mean_cosine_distance", "sampled_mean_silhouette"
  )
]
quality_display$largest_topic_share <- format_percent(
  quality_display$largest_topic_share
)
numeric_columns <- c(
  "normalized_size_entropy", "mean_cosine_distance",
  "sampled_mean_silhouette"
)
quality_display[numeric_columns] <- lapply(
  quality_display[numeric_columns],
  format_metric
)
knitr::kable(quality_display, row.names = FALSE)
```

## What the reduced `text` topics contain

```{r reduced-summary}
reduced_summary <- topic_summaries[
  topic_summaries$method == "text_reduced_6",
  c(
    "topic", "n_rows", "proportion_of_modeled", "top_terms",
    "representative_1", "representative_2"
  )
]
reduced_summary$proportion_of_modeled <- format_percent(
  reduced_summary$proportion_of_modeled
)
knitr::kable(reduced_summary, row.names = FALSE)
```

The dominant reduced topic mixes selection instructions, spatial-relation
questions, and quantity/box prompts. Smaller topics isolate recognizable narrow
templates—animal/person descriptions, number trains, retry prompts, grass
definitions, and untranslated Cyrillic text—but that does not rescue the overall
six-topic balance.

## Native `text` result

The ten largest non-outlier native topics are shown below. Their small sizes and
object/template-specific labels illustrate over-fragmentation rather than a
six-theme structure.

```{r native-topics}
native_display <- comparison$native_topic_info
native_display <- native_display[native_display$Topic >= 0, , drop = FALSE]
native_display <- native_display[order(-native_display$Count), , drop = FALSE]
native_display <- head(native_display, 10L)
knitr::kable(
  native_display[, c("Topic", "Count", "Name", "Representation")],
  row.names = FALSE
)
```

## Conclusion and recommendation

The confusing result is primarily a **corpus-definition problem**, not an SBERT
implementation problem. These translations contain many near-identical
instruction templates whose object names vary. Sentence embeddings naturally
encode both the template and its objects; native BERTopic then separates many
object/template variants, while forcing six clusters merges unrelated intents.

For six interpretable pedagogical topics, the next defensible analysis should
define the intended level of meaning first and then normalize the text—for
example, mask object names, numbers, colors, and positions before embedding—or
build a small reviewed intent taxonomy and evaluate clusters against it. Merely
switching from the Python-free `sbert` implementation to `text` does not solve
the issue.

## Reproducibility notes

```{r details}
knitr::kable(comparison$method_details, row.names = FALSE)
```
