---
title: "Thirty Semantic Topics in Computing Education Research — an sbert Tutorial"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
suppressMessages(devtools::load_all("..", quiet = TRUE))
fig_dir <- file.path("..", "tmp", "vignette_mcse_clause_figs")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# Why this tutorial

The MCSE corpus holds 16,453 bibliographic records of computing education
research. A field of that size cannot be read; it has to be mapped, and the
map has to be defensible — reviewers will ask why *these* topics and not
others. This tutorial repeats the deterministic workflow of the Levebee
feedback tutorial on a much harder corpus: long academic abstracts instead
of one-line messages, thirty topics instead of six, and a model chosen for
its input length rather than taken off the shelf.

Thirty is a *review granularity*, not a discovered truth: label-invariant
agreement between alternative runs of models this size is moderate, so the
partition should be read as one auditable organization of the field at a
chosen resolution. What makes it usable is that every step below is exactly
reproducible, and every topic ships with its own evidence: two keyword views
and the abstracts nearest its centroid.

| Component | Verb | Question it answers |
|---|---|---|
| Topic model | `topics()` | What areas exist, and how big are they? |
| Evaluation | `coherence()`, `topic_diversity()` | Which topics are trustworthy? |
| Frequent keywords | `terms(sort_by = "beta")` | What does each area talk about most? |
| Distinctive keywords | `topic_model$terms` | What does it talk about that others do not? |
| Evidence | `segment()` + `predict()` | Which real sentences support the label? |

```{r library, eval=FALSE}
library(sbert)
```

# The corpus: titles plus abstracts

Titles carry the claim, abstracts carry the content; embedding them together
gives each paper one vector that reflects both. Records with no usable
abstract (missing, or the literal "[NO ABSTRACT AVAILABLE]" placeholder) are
dropped, and the corpus is deduplicated — the same paper indexed twice
should vote once.

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
corpus <- dedupe(paste(trimws(usable$TI), trimws(usable$AB), sep = " — "))
nrow(corpus)
```

# Embedding long documents

Model choice is a length decision here. The default MiniLM truncates at 256
word pieces — fine for feedback messages, lossy for 139-word abstracts with
titles attached. `multi-qa-MiniLM-L6-cos-v1` keeps the same speed and
384-dimensional space but accepts 512 word pieces, which covers almost every
title-plus-abstract in this corpus without truncation
(`models()` lists the menu, including two 8,192-token models for even
longer documents). Encoding 15,551 documents is real work, so the model is
loaded once with eight threads; the run takes a few minutes and is
bit-identical every time.

```{r embeddings}
qa_minilm <- load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
embeddings <- encode(corpus$text, model = qa_minilm)
```

# The thirty-topic model

```{r model}
topic_model <- topics(
  corpus$text,
  n_topics = 30,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 3
)
topic_model
```

## Should you trust these topics?

With thirty topics the evaluation table is not a formality — it is the
reading order. NPMI coherence rewards topics whose top terms genuinely
co-occur in their abstracts; the topics at the top of this table have a
vocabulary of their own, and the ones at the bottom should be interpreted
from their representative abstracts, not their keywords.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
coherence(topic_model, measure = "npmi")
```

## Size on two scales

Unlike the Levebee corpus, this one barely repeats (15,660 usable records,
15,551 distinct texts), so the distinct and weighted shares nearly coincide
— the interesting quantity is plain size. A handful of large topics
covering the field's core and a long tail of specialized areas is the
expected shape for a research literature.

```{r sizes}
plot(topic_model, type = "sizes")
```

```{r weighted}
topic_sizes(topic_model, weights = corpus$n)
```

# Two keyword views, one topic

`terms(sort_by = "beta")` is the generative view — the words a topic uses most, which
in a single-discipline corpus is dominated by the field's shared vocabulary
("students", "programming", "course"). The class-based TF-IDF scores in
`topic_model$terms` are the discriminative view — the words a topic owns.
In a corpus like this the two lists disagree more than they did for the
feedback messages, and that disagreement is the point: the counts tell you
the topic belongs to computing education; only the TF-IDF terms tell you
which part of it.

```{r beta}
beta <- terms(topic_model, n = NULL, sort_by = "beta")
head(subset(beta, topic == 1), 8)
```

```{r terms}
head(subset(topic_model$terms, topic == 1), 8)
```

# Representative sentences, not papers

A whole abstract is poor evidence for a topic label: it buries the one
claim that matters under methods and boilerplate. The evidence should be
single statements — and *which* statements matters as much as how they are
cut. `segment()` splits every abstract at clause level (sentence ends
plus colons, semicolons, and subordinating hinges), so structured-abstract
headers ("CONTEXT:", "RESULTS:") stand apart and stacked clauses do not
glue into paragraphs. `representatives()` then chooses each topic's
evidence by *distinctiveness margin* rather than raw closeness: the nearest
unit to a centroid is systematically the longest one (centroids are
averages of whole abstracts, and long units resemble whole abstracts),
whereas the unit with the widest margin between its own topic and the
second-best is the statement that is unambiguously about this topic and no
other — naturally short and pointed.

```{r sentences}
abstract_sentences <- segment(unique(trimws(usable$AB)), merge_below = 5)
sentence_embeddings <- encode(abstract_sentences$text, model = qa_minilm)
evidence <- representatives(
  topic_model,
  abstract_sentences$text,
  embeddings = sentence_embeddings
)
```

# The topics, panel by panel

Each panel pairs the two keyword bars with the topic's three
highest-margin statements, shown verbatim — exactly the strings that were
embedded — with the distance to the topic centroid in parentheses. The
statements are the ground truth: the bars summarize, the statements
verify.

```{r panel-helper, include=FALSE}
palette_colors <- topic_palette(nrow(topic_model$topics))

draw_topic_panel <- function(model, beta, topic_id, color, path) {
  count_rows <- head(subset(beta, topic == topic_id), 8)
  score_rows <- head(subset(model$terms, topic == topic_id), 8)
  png(path, width = 2100, height = 1100, res = 300)
  op <- par(
    mfrow = c(1, 2), mar = c(2.6, 7.2, 2.2, 1.2),
    mgp = c(1.5, 0.35, 0), tcl = -0.2, family = "sans"
  )
  on.exit({par(op); dev.off()}, add = TRUE)
  draw_bar_column(
    values = rev(count_rows$count),
    labels = rev(count_rows$term),
    title = "Within-topic count",
    color = color,
    value_format = "%d"
  )
  draw_bar_column(
    values = rev(score_rows$score),
    labels = rev(score_rows$term),
    title = "Class-based TF-IDF",
    color = grDevices::adjustcolor(color, alpha.f = 0.55),
    value_format = "%.3f"
  )
  invisible(path)
}

draw_bar_column <- function(values, labels, title, color, value_format) {
  positions <- barplot(
    values,
    names.arg = labels,
    horiz = TRUE,
    las = 1,
    border = NA,
    col = color,
    xlim = c(0, max(values) * 1.22),
    cex.names = 0.72,
    cex.axis = 0.6,
    axes = FALSE
  )
  text(
    x = values,
    y = positions,
    labels = sprintf(value_format, values),
    pos = 4,
    cex = 0.6,
    col = "grey30"
  )
  title(main = title, adj = 0, cex.main = 0.85, font.main = 2, line = 0.6)
}
```

```{r panels, echo=FALSE, results="asis"}
invisible(lapply(
  seq_len(nrow(topic_model$topics)),
  function(topic_id) {
    topic_row <- subset(topic_model$topics, topic == topic_id)
    cat(sprintf(
      "# Topic %d — %s\n\n%d papers (%.1f%% of the corpus)\n\n",
      topic_id,
      topic_row$label,
      topic_row$n_documents,
      100 * topic_row$proportion
    ))
    figure_path <- file.path(fig_dir, sprintf("topic_%02d.png", topic_id))
    draw_topic_panel(topic_model, beta, topic_id, palette_colors[[topic_id]], figure_path)
    cat(sprintf("<img src=\"%s\" width=\"100%%\"/>\n\n", knitr::image_uri(figure_path)))
    statements <- subset(evidence, topic == topic_id)
    cat("**Representative statements**\n\n")
    cat(sprintf(
      "> %s *(%.3f)*\n\n",
      statements$text,
      statements$distance
    ), sep = "")
  }
))
```

# Where to go from here

`predict()` places any new abstract into this map without refitting;
`topic_membership()` grades papers that sit between areas;
`topic_gamma()` with `segment()` shows which individual sentences of
an abstract belong to which topic. Two refinements are worth knowing for a
publication-grade version: removing the corpus-wide search vocabulary from
term extraction (it never affects the clustering, only the labels), and
sweeping `n_topics` across candidate values with `coherence()` to
justify the chosen granularity.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
