---
title: "Thirty Sentence-Level Topics in Computing Education Research"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
suppressMessages(devtools::load_all("..", quiet = TRUE))
fig_dir <- file.path("..", "tmp", "vignette_mcse_sentence_figs")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# Why this tutorial

This is the sentence-level member of the MCSE vignette family, and the
level is not cosmetic: the *sentence is the unit of analysis*. Abstracts
are segmented first, the model is fitted on the sentence embeddings, the
topic centroids are averages of sentences, and every representative
statement shown under a topic is one of the units the model was trained
on — verbatim, in the exact form it was embedded. Nothing is quoted that
the model has not seen.

That changes what a topic *is*. The clause-level vignette maps papers:
each abstract votes once and topics are areas of the literature. Here
every sentence votes, so a methods sentence and a findings sentence from
the same paper can land in different topics — the model discovers the
field's recurring *statements* (what gets claimed, measured, and reported)
rather than its paper-level areas. The two maps are complementary and not
comparable topic-by-topic.

| Component | Verb | Question it answers |
|---|---|---|
| Topic model | `sbert_topics()` | What recurring statements exist, and how common are they? |
| Evaluation | `sbert_coherence()`, `sbert_diversity()` | Which topics are trustworthy? |
| Frequent keywords | `sbert_beta()` | What does each topic talk about most? |
| Distinctive keywords | `topic_model$terms` | What does it talk about that others do not? |
| Evidence | `sbert_representatives()` | Which training sentences define it? |

```{r library, eval=FALSE}
library(sbert)
```

# The corpus: abstract sentences

Records with no usable abstract are dropped, each remaining abstract is
segmented at sentence level (`merge_below = 5` folds fragments under five
words into their neighbor), and the sentences are deduplicated — academic
boilerplate repeats heavily, and a sentence that appears in forty
abstracts should vote once, with its multiplicity kept as a weight.
Titles are left out: they are compressed noun phrases, not sentences, and
would cluster by register instead of content.

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
sentences <- sbert_segment(
  unique(trimws(usable$AB)),
  level = "sentence",
  merge_below = 5
)
corpus <- sbert_dedupe(sentences$text)
nrow(corpus)
```

# Embedding the sentences

Sentences are short, so the 512-word-piece window of
`multi-qa-MiniLM-L6-cos-v1` is never a constraint here — the model is
chosen to keep this vignette in the same embedding space as its clause
and phrase siblings. Encoding roughly a hundred thousand sentences is the
expensive step; it runs once, on eight threads, and is bit-identical
every time.

```{r embeddings}
qa_minilm <- sbert_load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
embeddings <- sbert_encode(corpus$text, model = qa_minilm)
```

# The thirty-topic model

The model is fitted on the sentence embeddings themselves — thirty
centroids in sentence space.

```{r model}
topic_model <- sbert_topics(
  corpus$text,
  n_topics = 30,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 3
)
topic_model
```

## Should you trust these topics?

With thirty topics the evaluation table is not a formality — it is the
reading order. NPMI coherence rewards topics whose top terms genuinely
co-occur in their sentences; the topics at the top of this table have a
vocabulary of their own, and the ones at the bottom should be interpreted
from their representative sentences, not their keywords.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
sbert_coherence(topic_model, measure = "npmi")
```

## Size on two scales

Sentence corpora repeat where document corpora do not, so the two scales
genuinely differ here: the distinct count says how many different
sentences a topic owns, while the weighted share counts every occurrence
— a topic built on boilerplate ("data were collected through a survey")
grows on the second scale and not the first. Reading the two together
separates topics that are *varied* from topics that are *frequent*.

```{r sizes}
plot(topic_model, type = "sizes")
```

```{r weighted}
sbert_topic_sizes(topic_model, weights = corpus$n)
```

# Two keyword views, one topic

`sbert_beta()` is the generative view — the words a topic uses most,
dominated by the field's shared vocabulary. The class-based TF-IDF scores
in `topic_model$terms` are the discriminative view — the words a topic
owns. At sentence level the discriminative list is usually sharper than
in the document-level vignettes: sentences are single-minded, so a
topic's own vocabulary is less diluted by the rest of the abstract it
came from.

```{r beta}
beta <- sbert_beta(topic_model)
head(subset(beta, topic == 1), 8)
```

```{r terms}
head(subset(topic_model$terms, topic == 1), 8)
```

# Representative sentences from the training set

No second corpus and no second encoding: the candidates are exactly the
units the model was fitted on, under the embeddings it was fitted with.
`sbert_representatives()` ranks them by *distinctiveness margin* — the
gap between a sentence's own topic and the second-best — so the evidence
is the statement that belongs to this topic and could belong to no other,
not merely the one nearest a centroid.

```{r representatives}
evidence <- sbert_representatives(
  topic_model,
  corpus$text,
  embeddings = embeddings
)
```

# The topics, panel by panel

Each panel pairs the two keyword bars with the topic's three
highest-margin training sentences, verbatim, with the distance to the
topic centroid in parentheses. The sentences are the ground truth: the
bars summarize, the sentences verify.

```{r panel-helper, include=FALSE}
palette_colors <- sbert_palette(nrow(topic_model$topics))

draw_topic_panel <- function(model, beta, topic_id, color, path) {
  count_rows <- head(subset(beta, topic == topic_id), 8)
  score_rows <- head(subset(model$terms, topic == topic_id), 8)
  png(path, width = 2100, height = 1100, res = 300)
  op <- par(
    mfrow = c(1, 2), mar = c(2.6, 7.2, 2.2, 1.2),
    mgp = c(1.5, 0.35, 0), tcl = -0.2, family = "sans"
  )
  on.exit({par(op); dev.off()}, add = TRUE)
  draw_bar_column(
    values = rev(count_rows$count),
    labels = rev(count_rows$term),
    title = "Within-topic count",
    color = color,
    value_format = "%d"
  )
  draw_bar_column(
    values = rev(score_rows$score),
    labels = rev(score_rows$term),
    title = "Class-based TF-IDF",
    color = grDevices::adjustcolor(color, alpha.f = 0.55),
    value_format = "%.3f"
  )
  invisible(path)
}

draw_bar_column <- function(values, labels, title, color, value_format) {
  positions <- barplot(
    values,
    names.arg = labels,
    horiz = TRUE,
    las = 1,
    border = NA,
    col = color,
    xlim = c(0, max(values) * 1.22),
    cex.names = 0.72,
    cex.axis = 0.6,
    axes = FALSE
  )
  text(
    x = values,
    y = positions,
    labels = sprintf(value_format, values),
    pos = 4,
    cex = 0.6,
    col = "grey30"
  )
  title(main = title, adj = 0, cex.main = 0.85, font.main = 2, line = 0.6)
}
```

```{r panels, echo=FALSE, results="asis"}
invisible(lapply(
  seq_len(nrow(topic_model$topics)),
  function(topic_id) {
    topic_row <- subset(topic_model$topics, topic == topic_id)
    cat(sprintf(
      "# Topic %d — %s\n\n%d sentences (%.1f%% of the corpus)\n\n",
      topic_id,
      topic_row$label,
      topic_row$n_documents,
      100 * topic_row$proportion
    ))
    figure_path <- file.path(fig_dir, sprintf("topic_%02d.png", topic_id))
    draw_topic_panel(topic_model, beta, topic_id, palette_colors[[topic_id]], figure_path)
    cat(sprintf("<img src=\"%s\" width=\"100%%\"/>\n\n", knitr::image_uri(figure_path)))
    statements <- subset(evidence, topic == topic_id)
    cat("**Representative statements**\n\n")
    cat(sprintf(
      "> %s *(%.3f)*\n\n",
      statements$text,
      statements$distance
    ), sep = "")
  }
))
```

# Where to go from here

`predict()` places any new sentence into this map without refitting;
`sbert_membership()` grades sentences that sit between topics; and
running `predict()` on the segmented sentences of a *new* abstract
profiles that paper as a mixture over these statement topics. To compare
this map against the paper-level view of the same literature, read the
clause-level vignette — same corpus, same embedding model, different unit
of analysis.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
