---
title: "Thirty Context-Blended Topics in Computing Education Research"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
suppressMessages(devtools::load_all("..", quiet = TRUE))
fig_dir <- file.path("..", "tmp", "vignette_mcse_blended_figs")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# Why this tutorial

The sentence-level vignette fits topics on sentences alone; the clause-level
vignette fits on whole papers. This vignette mixes the two embeddings — the
right way. Naively pooling sentence units and document units into one
training set is unsound: units of different lengths occupy systematically
different regions of the embedding space (length-induced embedding
collapse, arXiv:2410.24200), handing the clustering a granularity axis to
cut along, and every sentence is a substring of its document, so the same
content would vote twice. `blend()` mixes *per unit* instead: each
sentence keeps what it says beyond its abstract — its context-orthogonal
residual — and inherits the rest of its direction from the abstract's
embedding, renormalized to unit length.

The practical payoff is twofold. A sentence that is ambiguous in isolation
("the control group scored higher") embeds near its abstract's subject
instead of floating between topics. And identical boilerplate sentences
from different abstracts now get *different* vectors, so recurring
methodological stock phrases cannot congeal into an artificial
boilerplate topic — each follows its own paper's subject. The unit of
analysis is still the sentence, and every statement quoted below is a
training unit, verbatim.

| Component | Verb | Question it answers |
|---|---|---|
| Context blending | `blend()` | How do sentences embed *in their paper's context*? |
| Topic model | `topics()` | What recurring statements exist, and how common are they? |
| Evaluation | `coherence()`, `topic_diversity()` | Which topics are trustworthy? |
| Frequent keywords | `terms(sort_by = "beta")` | What does each topic talk about most? |
| Distinctive keywords | `topic_model$terms` | What does it talk about that others do not? |
| Evidence | `representatives()` | Which training sentences define it? |

```{r library, eval=FALSE}
library(sbert)
```

# The corpus: abstract sentences with their abstracts

Records with no usable abstract are dropped and each remaining abstract is
segmented at sentence level (`merge_below = 5` folds fragments under five
words into their neighbor). Unlike the pure sentence-level vignette, the
sentences are *not* deduplicated: under blending, the same sentence in two
different abstracts is two different pieces of evidence with two different
vectors, and collapsing them would throw the context away.

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
abstracts <- unique(trimws(usable$AB))
sentences <- segment(abstracts, level = "sentence", merge_below = 5)
nrow(sentences)
```

# Blending the two embeddings

One verb does the mixing: it encodes the sentences, encodes the abstracts,
and blends each sentence's context-orthogonal residual with its parent
abstract's direction. `alpha = 0.5` gives equal weight to what the
sentence alone says and to where its abstract sits; `alpha = 1` would keep
only the sentence's own contribution, `alpha = 0` would collapse every
sentence onto its abstract. This is the expensive step — two full encoding
passes — and it is bit-identical on every run.

```{r embeddings}
qa_minilm <- load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
embeddings <- blend(
  sentences,
  abstracts,
  model = qa_minilm,
  alpha = 0.5
)
```

# The thirty-topic model

The model is fitted on the blended embeddings — thirty centroids over
sentences-in-context.

```{r model}
topic_model <- topics(
  sentences$text,
  n_topics = 30,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 3
)
topic_model
```

## Should you trust these topics?

The same reading order as in the sibling vignettes: NPMI coherence first.
Compare this table against the pure sentence-level vignette — that
comparison, at identical corpus and topic count, is the empirical test of
whether context blending helps on this corpus, and published evidence for
blending comes from retrieval, not topic modeling, so the answer is not a
foregone conclusion.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
coherence(topic_model, measure = "npmi")
```

## Topic sizes

Sizes count sentences: with blending active there is no deduplication, so
every occurrence of a repeated sentence counts where its own abstract
placed it.

```{r sizes}
plot(topic_model, type = "sizes")
```

```{r weighted}
topic_sizes(topic_model)
```

# Two keyword views, one topic

`terms(sort_by = "beta")` is the generative view — the words a topic uses most. The
class-based TF-IDF scores in `topic_model$terms` are the discriminative
view — the words a topic owns. Blending typically moves these lists closer
to the *document-level* vocabulary than the pure sentence fit does: each
sentence drags half of its abstract's direction with it, so topics
inherit paper-level subject terms alongside sentence-level phrasing.

```{r beta}
beta <- terms(topic_model, n = NULL, sort_by = "beta")
head(subset(beta, topic == 1), 8)
```

```{r terms}
head(subset(topic_model$terms, topic == 1), 8)
```

# Representative sentences from the training set

The candidates are exactly the units the model was fitted on, under the
blended embeddings it was fitted with. `representatives()` ranks
them by *distinctiveness margin* — the gap between a sentence's own topic
and the second-best — so the evidence is the statement that belongs to
this topic and could belong to no other.

```{r representatives}
evidence <- representatives(
  topic_model,
  sentences$text,
  embeddings = embeddings
)
```

# The topics, panel by panel

Each panel pairs the two keyword bars with the topic's three
highest-margin training sentences, verbatim, with the distance to the
topic centroid (in the blended space) in parentheses. The sentences are
the ground truth: the bars summarize, the sentences verify.

```{r panel-helper, include=FALSE}
palette_colors <- topic_palette(nrow(topic_model$topics))

draw_topic_panel <- function(model, beta, topic_id, color, path) {
  count_rows <- head(subset(beta, topic == topic_id), 8)
  score_rows <- head(subset(model$terms, topic == topic_id), 8)
  png(path, width = 2100, height = 1100, res = 300)
  op <- par(
    mfrow = c(1, 2), mar = c(2.6, 7.2, 2.2, 1.2),
    mgp = c(1.5, 0.35, 0), tcl = -0.2, family = "sans"
  )
  on.exit({par(op); dev.off()}, add = TRUE)
  draw_bar_column(
    values = rev(count_rows$count),
    labels = rev(count_rows$term),
    title = "Within-topic count",
    color = color,
    value_format = "%d"
  )
  draw_bar_column(
    values = rev(score_rows$score),
    labels = rev(score_rows$term),
    title = "Class-based TF-IDF",
    color = grDevices::adjustcolor(color, alpha.f = 0.55),
    value_format = "%.3f"
  )
  invisible(path)
}

draw_bar_column <- function(values, labels, title, color, value_format) {
  positions <- barplot(
    values,
    names.arg = labels,
    horiz = TRUE,
    las = 1,
    border = NA,
    col = color,
    xlim = c(0, max(values) * 1.22),
    cex.names = 0.72,
    cex.axis = 0.6,
    axes = FALSE
  )
  text(
    x = values,
    y = positions,
    labels = sprintf(value_format, values),
    pos = 4,
    cex = 0.6,
    col = "grey30"
  )
  title(main = title, adj = 0, cex.main = 0.85, font.main = 2, line = 0.6)
}
```

```{r panels, echo=FALSE, results="asis"}
invisible(lapply(
  seq_len(nrow(topic_model$topics)),
  function(topic_id) {
    topic_row <- subset(topic_model$topics, topic == topic_id)
    cat(sprintf(
      "# Topic %d — %s\n\n%d sentences (%.1f%% of the corpus)\n\n",
      topic_id,
      topic_row$label,
      topic_row$n_documents,
      100 * topic_row$proportion
    ))
    figure_path <- file.path(fig_dir, sprintf("topic_%02d.png", topic_id))
    draw_topic_panel(topic_model, beta, topic_id, palette_colors[[topic_id]], figure_path)
    cat(sprintf("<img src=\"%s\" width=\"100%%\"/>\n\n", knitr::image_uri(figure_path)))
    statements <- subset(evidence, topic == topic_id)
    cat("**Representative statements**\n\n")
    cat(sprintf(
      "> %s *(%.3f)*\n\n",
      statements$text,
      statements$distance
    ), sep = "")
  }
))
```

# Where to go from here

`alpha` is the experiment: sweep it (0.25, 0.5, 0.75) with
`coherence()` at fixed `n_topics` to see how much context this
corpus wants — the blend at `alpha = 1` approaches a context-orthogonal
fit, at `alpha = 0` a paper-level fit read through sentences. To place a
new abstract into this map, segment it, blend against its own text, and
`predict()`. The pure sentence-level and clause-level vignettes are the
baselines this design should be judged against.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
