---
title: "Thirty Phrase-Level Topics in Computing Education Research"
author: "sbert 0.4.0"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 1
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
suppressMessages(devtools::load_all("..", quiet = TRUE))
fig_dir <- file.path("..", "tmp", "vignette_mcse_phrase_figs")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
mcse_path <- file.path(
  "/Users/mohammedsaqr/Library/CloudStorage/GoogleDrive-saqr@saqr.me",
  "Other computers/My MacBook Pro (1)/AKA/Networks_CSE/MCSE_gold.RDS"
)
```

# Why this tutorial

This is the phrase-level member of the MCSE vignette family, and the
level is not cosmetic: the *phrase is the unit of analysis*. Abstracts
are segmented at the finest granularity the package offers — sentence
ends, clause separators, and commas — the model is fitted on the phrase
embeddings, the topic centroids are averages of phrases, and every
representative unit shown under a topic is one of the units the model
was trained on, verbatim, in the exact form it was embedded.

Phrases are the field's smallest reusable expressions: named methods,
measured constructs, list items, qualifiers. A phrase-level model
therefore clusters *terminology in context* rather than papers or
claims — expect tighter, more lexical topics than in the sentence-level
sibling, and expect some units to be fragments rather than checkable
claims. That trade is the point of this level; the `merge_below = 5`
guard keeps the fragments from degenerating into single words.

| Component | Verb | Question it answers |
|---|---|---|
| Topic model | `topics()` | What recurring expressions exist, and how common are they? |
| Evaluation | `coherence()`, `topic_diversity()` | Which topics are trustworthy? |
| Frequent keywords | `terms(sort_by = "beta")` | What does each topic talk about most? |
| Distinctive keywords | `topic_model$terms` | What does it talk about that others do not? |
| Evidence | `representatives()` | Which training phrases define it? |

```{r library, eval=FALSE}
library(sbert)
```

# The corpus: abstract phrases

Records with no usable abstract are dropped, each remaining abstract is
segmented at phrase level (`merge_below = 5` folds shards under five
words into their neighbor), and the phrases are deduplicated — short
units repeat far more than sentences do, so multiplicity is kept as a
weight and matters throughout. Titles are left out: they are compressed
noun phrases of a different register.

```{r data}
mcse <- readRDS(mcse_path)
usable <- subset(mcse, !is.na(AB) & !startsWith(trimws(AB), "[NO ABSTRACT"))
phrases <- segment(
  unique(trimws(usable$AB)),
  level = "phrase",
  merge_below = 5
)
corpus <- dedupe(phrases$text)
nrow(corpus)
```

# Embedding the phrases

Phrases are the shortest units in the family, so the 512-word-piece
window of `multi-qa-MiniLM-L6-cos-v1` is never a constraint — the model
is chosen to keep this vignette in the same embedding space as its
sentence and clause siblings. This level has the largest corpus of the
three; encoding it is the expensive step, runs once on eight threads,
and is bit-identical every time.

```{r embeddings}
qa_minilm <- load_model("multi-qa-MiniLM-L6-cos-v1", threads = 8L)
embeddings <- encode(corpus$text, model = qa_minilm)
```

# The thirty-topic model

The model is fitted on the phrase embeddings themselves — thirty
centroids in phrase space.

```{r model}
topic_model <- topics(
  corpus$text,
  n_topics = 30,
  embeddings = embeddings,
  n_terms = 8,
  n_representatives = 3
)
topic_model
```

## Should you trust these topics?

With thirty topics the evaluation table is not a formality — it is the
reading order. NPMI coherence rewards topics whose top terms genuinely
co-occur in their phrases; the topics at the top of this table have a
vocabulary of their own, and the ones at the bottom should be
interpreted from their representative phrases, not their keywords.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
coherence(topic_model, measure = "npmi")
```

## Size on two scales

At phrase level the two scales diverge more than anywhere else in the
family: formulaic expressions repeat massively, so a topic can be small
in distinct phrases yet large in occurrences. The distinct count
measures a topic's variety; the weighted share measures its presence in
the literature. Topics whose weighted share far exceeds their distinct
share are built on the field's stock phrases.

```{r sizes}
plot(topic_model, type = "sizes")
```

```{r weighted}
topic_sizes(topic_model, weights = corpus$n)
```

# Two keyword views, one topic

`terms(sort_by = "beta")` is the generative view — the words a topic uses most.
The class-based TF-IDF scores in `topic_model$terms` are the
discriminative view — the words a topic owns. At phrase level the two
views converge more than at any other level: a phrase is often little
more than its keywords, so the bars and the units below them should
visibly agree. Where they do not, the topic is held together by
embedding similarity that the vocabulary does not show — read its
phrases.

```{r beta}
beta <- terms(topic_model, n = NULL, sort_by = "beta")
head(subset(beta, topic == 1), 8)
```

```{r terms}
head(subset(topic_model$terms, topic == 1), 8)
```

# Representative phrases from the training set

No second corpus and no second encoding: the candidates are exactly the
units the model was fitted on, under the embeddings it was fitted with.
`representatives()` ranks them by *distinctiveness margin* — the
gap between a phrase's own topic and the second-best — so the evidence
is the expression that belongs to this topic and could belong to no
other, not merely the one nearest a centroid.

```{r representatives}
evidence <- representatives(
  topic_model,
  corpus$text,
  embeddings = embeddings
)
```

# The topics, panel by panel

Each panel pairs the two keyword bars with the topic's three
highest-margin training phrases, verbatim, with the distance to the
topic centroid in parentheses. The phrases are the ground truth: the
bars summarize, the phrases verify.

```{r panel-helper, include=FALSE}
palette_colors <- topic_palette(nrow(topic_model$topics))

draw_topic_panel <- function(model, beta, topic_id, color, path) {
  count_rows <- head(subset(beta, topic == topic_id), 8)
  score_rows <- head(subset(model$terms, topic == topic_id), 8)
  png(path, width = 2100, height = 1100, res = 300)
  op <- par(
    mfrow = c(1, 2), mar = c(2.6, 7.2, 2.2, 1.2),
    mgp = c(1.5, 0.35, 0), tcl = -0.2, family = "sans"
  )
  on.exit({par(op); dev.off()}, add = TRUE)
  draw_bar_column(
    values = rev(count_rows$count),
    labels = rev(count_rows$term),
    title = "Within-topic count",
    color = color,
    value_format = "%d"
  )
  draw_bar_column(
    values = rev(score_rows$score),
    labels = rev(score_rows$term),
    title = "Class-based TF-IDF",
    color = grDevices::adjustcolor(color, alpha.f = 0.55),
    value_format = "%.3f"
  )
  invisible(path)
}

draw_bar_column <- function(values, labels, title, color, value_format) {
  positions <- barplot(
    values,
    names.arg = labels,
    horiz = TRUE,
    las = 1,
    border = NA,
    col = color,
    xlim = c(0, max(values) * 1.22),
    cex.names = 0.72,
    cex.axis = 0.6,
    axes = FALSE
  )
  text(
    x = values,
    y = positions,
    labels = sprintf(value_format, values),
    pos = 4,
    cex = 0.6,
    col = "grey30"
  )
  title(main = title, adj = 0, cex.main = 0.85, font.main = 2, line = 0.6)
}
```

```{r panels, echo=FALSE, results="asis"}
invisible(lapply(
  seq_len(nrow(topic_model$topics)),
  function(topic_id) {
    topic_row <- subset(topic_model$topics, topic == topic_id)
    cat(sprintf(
      "# Topic %d — %s\n\n%d phrases (%.1f%% of the corpus)\n\n",
      topic_id,
      topic_row$label,
      topic_row$n_documents,
      100 * topic_row$proportion
    ))
    figure_path <- file.path(fig_dir, sprintf("topic_%02d.png", topic_id))
    draw_topic_panel(topic_model, beta, topic_id, palette_colors[[topic_id]], figure_path)
    cat(sprintf("<img src=\"%s\" width=\"100%%\"/>\n\n", knitr::image_uri(figure_path)))
    statements <- subset(evidence, topic == topic_id)
    cat("**Representative statements**\n\n")
    cat(sprintf(
      "> %s *(%.3f)*\n\n",
      statements$text,
      statements$distance
    ), sep = "")
  }
))
```

# Where to go from here

`predict()` places any new phrase into this map without refitting;
`topic_membership()` grades phrases that sit between topics; and running
`predict()` on the segmented phrases of a new abstract profiles that
paper as a mixture over these expression topics. To compare this map
against the sentence-level and paper-level views of the same literature,
read the sibling vignettes — same corpus, same embedding model,
different unit of analysis.

```{r session, echo=FALSE}
sessionInfo()$R.version$version.string
```
